### 2021-03-06 manuscript.Rmd
another day, another project

---
title: 'Project 3: Variant Calling'
author: "J Moggridge"
date: "06/03/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2021-03-13 manuscript.Rmd
much progress

@@
 ---
-title: 'Project 3: Variant Calling'
+title: 'Project 3: Variant calling in Burbot (*Lota lota*) using `freebayes` and `mpileup`'
 author: "J Moggridge"
 date: "06/03/2021"
-output: pdf_document
+output: 
+  pdf_document:
+    highlight: tango
+bibliography: references.bib
 ---
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+```{r setup, echo=F}
+knitr::opts_chunk$set(warning = T, message = T, eval = T, fig.align = 'center', echo = T, tidy = T, cache = T)
 ```
 
-## R Markdown
+```{=html}
+<!-- For this project, we will use multiple variant calling softwares to identify variable sites in genomic data.
+We will again use the burbot dataset that we used in Project 2, because this dataset is a nice size for teaching and learning (big enough to be realistic, not so large as to be unwieldy).
 
-This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.
+ From the three variant callers we used in class –GATK Haplotype Caller (McKenna et  al., 2010), Freebayes (Garrison & Marth, 2012), and bcftools mpileup (Liet al., 2009) – please select two to use for this project.
 
-When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
-You can embed an R code chunk like this:
+ Use the 10 individual sorted.bam files from Project 2 (along with indexes, .bai) for this analysis.
+You may choose to use your own results from Problem Set 2, or may use .bam files that I will provide.
 
-```{r cars}
-summary(cars)
+ Call variants for these 10 individuals using two different software packages. 
+ Filter resulting vcfs for quality, as shown in class, using what you think is a reasonable threshold. 
+ - check transition/transversion ratio: 
+ Compare the resulting vcf files in several ways. First, compare the number of SNPs and the overlap of these SNPs across files.
+
+ Then, plot:
+  - the minor allele frequency 
+  - and depth of all SNPs (summed across individuals), and discuss the variation across loci.
+
+ Compare across vcfs produced by different variant callers. 
+ ###
+ https://samtools.github.io/bcftools/howtos/index.html
+ -->
+```
+# Introduction
+
+Variant calling / genotyping Sequencing error is high, need to filter and assess likelihoods. Low concordance between methods, however. No ground truth available to ascertain exactly what is an error.
+
+Comparing BCFTOOLS mpileup and freebayes
+
+`mpileup` uses ... @Li2011
+
+`freebayes` uses... @garrison2012
+
+---
+
+# Methods
+
+<!-- (chmod 755 /scratch/username/directory/ -R). -->
+
+<!-- srun --pty --account="def-nricker" -t 0-03:00:00 --mem=32000 /bin/bash -->
+
+### Dataset
+
+I used data from project \#2 in this work, specifically the reads mapped to the burbot reference with the`bowtie2` aligner.
+I used `bcftools mpileup` and `GATK HaplotypeCaller` to call variants for the ten individuals using these alignments.
+
+I first needed to convert my `.sam` files to `.bam` format and index them, as well as adding read groups using `picard` (necessary for `freebayes`).
+
+```{r sam-to-bam and readgroups, engine = 'bash', eval = FALSE, echo=TRUE}
+#!/bin/bash
+## to convert .sam to .bam, then sort & index
+module load samtools
+for file in bam/*.sam; do
+  name=`echo $file | sed 's/\.sam//'`;
+  echo $name;
+  samtools view -b -S -o $name.bam $file;
+  samtools sort $name.bam -o $name.sorted.bam;
+  samtools index $name.sorted.bam
+done
+
+# add readgroups and re-index; show output to verify
+module load picard 
+group=1
+for file in bam/*.sorted.bam
+do
+  label=`echo $file | sed 's/.sorted.bam//'`
+  java -jar $EBROOTPICARD/picard.jar AddOrReplaceReadGroups I=$file \
+    O=$label.rg.sorted.bam RGID=$group RGLB=lib1 RGPL=illumina RGPU=unit1 RGSM=$label
+  samtools index $label.rg.sorted.bam
+  echo "group number: $group; file: $file; read group:"
+  samtools view -H $label.rg.sorted.bam | grep "@RG"
+  group=$((group+1))
+done
 ```
 
-## Including Plots
+### `bcftools mpileup/call` and `freebayes`
 
-You can also embed plots, for example:
+I ran `bcftools` variant calling for the 10 .sorted.bam files.
+I added arguments to provide annotations for read depth and allelic depth (-a DP,AD), to skip indels, to specify Illumina as the sequencing platform (-P), to consider up to 1000 reads at a locus instead of the default 250 (--max-depth).
+In variant calling, I specified to use the multi-allelic caller, to skip indels, and to provide the genotype quality scores (GQ).
+Bash commands given:
 
-```{r pressure, echo=FALSE}
-plot(pressure)
+```{r mpileup, engine = 'bash', eval = FALSE, echo=TRUE}
+module load nixpkgs/16.09  gcc/7.3.0 bcftools/1.9
+# mpileup and variant calling
+bcftools mpileup -Ou -a DP,AD --skip-indels -P ILLUMINA  --max-depth 1000 \
+-f ./burbot_ref/GCA_900302385.1_ASM90030238v1_genomic.fna ./bam/*.rg.sorted.bam | \
+bcftools call -mv --format-fields GQ -o  mpileup.call.vcf
 ```
 
-Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
+I called SNPs with `freebayes` with similar settings applied for `mpileup`.
+Unfortunately there seems to be no equivalent for maximum reads consider (--max-depth), though this probably doesn't matter too much.
+Bash commands given:
+
+```{r freebayes, engine = 'bash', eval = FALSE, echo = TRUE}
+module --force purge; module load nixpkgs/16.09  gcc/7.3.0 freebayes/1.2.0
+# create a list of files to pass to as inputs
+> files.txt
+for file in bam/*.rg.sorted.bam; do echo $file >> files.txt; done
+# free-bayes calling
+freebayes -f burbot_ref/GCA_900302385.1_ASM90030238v1_genomic.fna  \
+--genotype-qualities --bam-list files.txt | vcffilter -f "TYPE = snp" > freebayes.call.vcf
+module unload freebayes
+```
+
+## Filtering variant calls
+
+Biallelic SNPs were retained if they had a minimum quality score of 30, minimum genotyping quality score of 10, and been genotyped in at least 50% of individuals.
+Given the small number of individuals, I did not feel it necessary to filter by allele frequency.
+There was no need to exclude any individuals, as none had more that 10% missing data (from `--missing-indv`) .
+
+<!-- (--remove-filtered-all) is like  (--apply-filter "PASS")-->
+
+<!-- (--min-alleles 2 --max-alleles 2) for biallelic -->
+
+<!-- (--minQ 30 --minGQ 10 --max-missing 0.5) for QUAL, GQ, and prop missing data -->
+
+<!-- (-recode --recode-INFO-all) to get a new vcf with annotation data -->
+
+```{r filtering, engine='bash', eval = FALSE, echo = TRUE}
+module --force purge; module load StdEnv/2020 vcftools/0.1.16
+# filter mpileup calls
+vcftools --vcf mpileup.call.vcf --minQ 30 --minGQ 10 --max-missing 0.5 \
+--min-alleles 2 --max-alleles 2 --remove-indels \
+--remove-filtered-all --recode --recode-INFO-all --out mpileup.q30gq10g5
+# same thing for freebayes
+vcftools --vcf freebayes.call.vcf --minQ 30 --minGQ 10 --max-missing 0.5 \
+--min-alleles 2 --max-alleles 2 --remove-indels \
+--remove-filtered-all --recode --recode-INFO-all --out freebayes.q30gq10g5
+
+## check individuals; the amount of data missing is < 12% for all in each file
+vcftools --vcf mpileup.q30gq10g5.recode.vcf --missing-indv \
+  --out mpileup.q30gq10g5
+cat mpileup.q30gq10g5.imiss
+vcftools --vcf freebayes.q30gq10g5.recode.vcf --missing-indv \
+  --out freebayes.q30gq10g5
+cat freebayes.q30gq10g5.imiss
+```
+
+### Count SNPs with `bcftools stats` and `isec`
+
+```{r count snps, engine = 'bash', eval = FALSE, echo = TRUE}
+module load StdEnv/2020 bcftools/1.10.2
+# count the snps: 40,265 and 29,176
+bcftools stats mpileup.q30gq10g5.recode.vcf | grep "number of SNPs:" | cut -f4
+bcftools stats freebayes.q30gq10g5.recode.vcf | grep "number of SNPs:" | cut -f4 
+# no indels present, so I did one thing right at least
+bcftools stats mpileup.q30gq10g5.recode.vcf | grep "number of indels:"
+bcftools stats freebayes.q30gq10g5.recode.vcf | grep "number of indels:"
+
+# compare files to get intersection - doesnt work
+### vcftools --vcf mpileup.q30gq10g5.recode.vcf  --diff freebayes.q30gq10g5.recode.vcf \
+### --diff-site --out vcf_compare_vcftools
+```
+
+The differences in SNP sets were 
+```{r, engine = 'bash', eval = FALSE, echo = TRUE}
+# prepare mpileup snps by zipping and indexing
+bgzip mpileup.q30gq10g5.recode.vcf
+bcftools index mpileup.q30gq10g5.recode.vcf.gz
+# Prepare freebayes snps and indexing
+bgzip freebayes.q30gq10g5.recode.vcf
+bcftools index freebayes.q30gq10g5.recode.vcf.gz
+# Compute intersection of 2 files to isec/
+bcftools isec -p isec mpileup.q30gq10g5.recode.vcf.gz freebayes.q30gq10g5.recode.vcf.gz
+# Get this data into a useable format
+> Overlap.vals
+for file in isec/*.vcf
+do 
+  bcftools stats $file| grep "number of SNPs:" | cut -f4 >> Overlap.vals
+done
+echo $'unique to mpileup\nunique to freebayes\nshared\nshared' > Overlap.names
+paste  Overlap.names Overlap.vals > Overlap.tsv
+sed -i '1s/^/SNPs\tCount\n/' Overlap.tsv
+cat Overlap.tsv
+rm Overlap.names Overlap.vals
+# from local terminal
+# scp jmoggrid@cedar.computecanada.ca:/scratch/jmoggrid/Project3/Overlap.tsv .
+```
+
+```{r vcf allele freq, engine = 'bash', eval = FALSE, echo = TRUE}
+# allele frequency stats > output to .frq files
+vcftools --gzvcf mpileup.q30gq10g5.recode.vcf.gz --freq --out mpileup.q30gq10g5
+vcftools --gzvcf freebayes.q30gq10g5.recode.vcf.gz --freq --out freebayes.q30gq10g5
+# depth for each individual >
+vcftools --vcf file.vcf --depth -c > ind_depth.txt
+vcftools --gzvcf mpileup.q30gq10g5.recode.vcf.gz --depth -c > mpileup_ind_depth.txt
+vcftools --gzvcf freebayes.q30gq10g5.recode.vcf.gz --depth -c > freebayes_ind_depth.txt
+```
+
+<!-- quality, as shown in class, using what you think is a reasonable threshold.
+Compare the resulting vcf files in several ways.
+First, *compare the number of SNPs* and the *overlap of these SNPs across files*.
+-->
+
+<!-- Then, plot the *minor allele frequency* and depth of all SNPs (summed across individuals), and discuss the variation across loci. Compare across vcfs produced by different variant callers. -->
+
+---
+
+### R code
+
+#### Reading in the data:
+
+    # get data down from cedar in local terminal:
+    # scp jmoggrid@cedar.computecanada.ca:/scratch/jmoggrid/Project3/Overlap.tsv .
+    # scp jmoggrid@cedar.computecanada.ca:/scratch/jmoggrid/Project3/*.frq .
+    # scp jmoggrid@cedar.computecanada.ca:/scratch/jmoggrid/Project3/*ind_depth .
+
+```{r load data, message=FALSE}
+#! R code:
+library(tidyverse, quietly = T)
+# Overlap counts
+Overlap <- read_delim('./data/Overlap.tsv', '\t') %>%  distinct() %>% 
+  bind_rows(
+    tibble(SNPs = c('mpileup', 'freebayes'), 
+           Count = c(.[1,2][[1]] + .[3,2][[1]], .[2,2][[1]] + .[3,2][[1]])
+    )) %>% 
+  arrange(desc(Count))
+# Allele frequencies
+frq_names <- c('CHROM','POS', 'N_ALLELES','N_CHR','AL1','AL2')
+Allele_freq <- 
+  bind_rows(
+    read_tsv('./data/freebayes.q30gq10g5.frq', 
+             col_names = frq_names, skip = 1L) %>% 
+      mutate(Caller = 'freebayes'),
+    read_tsv('./data/mpileup.q30gq10g5.frq', 
+             col_names = frq_names, skip = 1L) %>% 
+      mutate(Caller = 'mpileup')
+  ) %>% 
+  mutate(G1.allele = str_remove(AL1, ':.+'),
+         G1.freq = str_remove(AL1, '.:'),
+         G2.allele = str_remove(AL2, ':.+'),
+         G2.freq = str_remove(AL2, '.:')
+        ) %>% 
+  mutate(across(contains('freq'), as.numeric))
+
+# Depth
+Depth <- bind_rows(
+  read_tsv('./data/mpileup_ind_depth.txt') %>% mutate(Caller = 'mpileup'),
+  read_tsv('./data/freebayes_ind_depth.txt') %>% mutate(Caller = 'freebayes')) %>% 
+  mutate(INDV = str_remove_all(INDV, 'bam.|.bowtie.burbot')) %>% 
+  arrange(INDV)
+Depth
+```
+
+---
+
+\newpage
+
+# Results
+
+```{r table snps and overlap, echo=FALSE}
+Overlap %>% 
+  pander::pander(caption = 'Total, shared, and unique SNPs called by mpileup and freebayes', justify='left')
+```
+
+```{r , message=FALSE}
+Depth %>% 
+  pivot_longer(cols = c(N_SITES, MEAN_DEPTH)) %>% 
+   ggplot(aes(y=INDV, x=value)) +
+  geom_point() +
+  geom_segment(aes(x = 0, xend=value, yend=INDV)) +
+  facet_wrap(Caller~name, ncol=2, scales='free_x') +
+  labs(y='', subtitle = 'Depth stats') +
+  theme_bw()
+```
+
+```{r}
+ggplot(Allele_freq %>% filter(G1.freq != 0), aes(x = G1.freq, fill=G1.allele)) +
+  geom_histogram(bins = 20, colour = 'white',) +
+  facet_wrap(~Caller) +
+  rcartocolor::scale_fill_carto_d() +
+  labs(x = 'minor allele frequency', fill = 'Allele') +
+  theme_bw()
+```
+
+```{r, fig.cap='Mean read depth of SNPs for each individual (points) and the distribution (line)'}
+set.seed(45)
+library(ggbeeswarm)
+ggplot(Depth, aes(y = MEAN_DEPTH, x = Caller)) +
+ geom_violin(alpha=0.1, draw_quantiles = TRUE, trim = F) +
+  # geom_jitter(aes(color=INDV), height = 0.12) +
+  ggbeeswarm::geom_quasirandom(aes(color=INDV), size=3, width = 0.3) + 
+  rcartocolor::scale_color_carto_d() +
+  guides(colour = guide_legend(override.aes = list(size=5, pch=15))) +
+  facet_wrap(~Caller, scales = 'free_x') +
+  labs(y = 'mean depth per locus', x = '', color = 'Individual') +
+  theme_bw() +  
+  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
+```
+
+\newpage
+
+# Discussion
+
+------------------------------------------------------------------------
+
+\newpages
+
+# References
+
+------------------------------------------------------------------------
+
+<!-- ## `bcftools` filter: multiple qualities -->
+
+<!-- module load bcftools -->
+
+<!-- # filter snps with qual 20 & depth ? -->
+
+<!-- touch tstv.txt -->
+
+<!-- echo "i,tstv" > tstv.txt -->
+
+<!-- for i in seq 10 10 50 -->
+
+<!-- do -->
+
+<!--   tstv=`bcftools filter -i'%QUAL>$i' calls.vcf.gz | bcftools stats > mpileup.qual_$i.stats -->
+
+<!--   grep TSTV mpileup.qual_$i.stats -->
+
+<!--   echo "$i,$tstv" >> tstv.txt -->
+
+<!-- done -->
+
+<!-- **do we need the multi-allelic caller -m, if we are only wanting biallelic snps?** -->
+
+<!-- **do we need format-fields GQ if we aren't filtering on GQ?** -->
+
+<!-- - `bcftools mpileup` options:     -->
+
+<!--  - `Ou`: to to work with uncompressed BCF output -->
+
+<!--  - `-a`: Annotate: -->
+
+<!--    - `AD`: allelic depth -->
+
+<!--    - `PD`: Number of high-quality bases (Number=1,Type=Integer) -->
+
+<!--  - `--skip-indels`: Do not perform INDEL calling -->
+
+<!--  - `-P`: Platform; used for calling indels... -->
+
+<!-- - `bcftools call` options:   -->
+
+<!--   - `-v, --variants-only`: output variant sites only -->
+
+<!--   - `-m, --multiallelic-caller`: alternative model for multiallelic and rare-variant calling designed to overcome known limitations in -c calling model (conflicts with -c) -->
+
+<!--   - `-c, --consensus-caller`: the original samtools/bcftools calling method (conflicts with -m) -->
+
+<!-- - `filter` ops:   -->
+
+<!--   - --set-GTs: set genotypes of failed samples to missing value (.) or reference allele (0)   -->
+
+<!--   - `--min-alleles 2 --max-alleles 2`: to only get biallelic SNPs   -->
+
+<!-- - `view` options: -->
+
+<!-- --apply-filter "PASS" -->
+
+<!-- -i 'INFO/SVNTYPE~"DEL"' -->
+
+<!-- ## GATK -->
+
+<!-- ```{r gatk calling, engine = 'bash'} -->
+
+<!-- module --force purge -->
+
+<!-- module load picard -->
+
+<!-- # create the reference index for GATK -->
+
+<!-- java -jar $EBROOTPICARD/picard.jar CreateSequenceDictionary \ -->
+
+<!--   R=burbot_ref/GCA_900302385.1_ASM90030238v1_genomic.fna \ -->
+
+<!--   O=burbot_ref/GCA_900302385.1_ASM90030238v1_genomic.dict -->
+
+<!-- module --force purge -->
+
+<!-- module load nixpkgs/16.09 gatk/4.0.0.0  -->
+
+<!-- ## Basic variant calling command;  -->
+
+<!-- ## bam files are passed from files.txt with ` -I ` before each name -->
+
+<!-- gatk --java-options "-Xmx4g" HaplotypeCaller \ -->
+
+<!-- -R burbot_ref/GCA_900302385.1_ASM90030238v1_genomic.fna \ -->
+
+<!-- `awk '{printf(" -I %s ",$0);}' files.txt` \ -->
+
+<!-- -O gatk.call.vcf -->
+
+<!-- ## generates a billion warning messages (takes many hours) -->
+
+<!-- ``` -->
+
+<!-- vcftools --vcf SMALL_bcftools_filtered.vcf --diff SMALL_freebayes.vcf --diff-site --out bcftools_v_freebayes -->
+
+<!-- ### other Considerations ???-->
+
+<!-- - Remove PCR duplicates? - alignment ... -->
+
+<!-- - Min-quality threshold for variants? try multiple? GATK defaults to 30? -->
+
+<!-- <!-- - Local realignment to improve accuracy for indels (callers favor mismatches) -->
+
+<!-- - transition/transversion ratio as a metric of quality? -->
+
+<!-- ## `bcftools` simple filtering -->
+
+<!-- From variant calls by `freebayes` and `mpileup`, biallelic SNPs (`--min-alleles 2 --max-alleles 2`, `--types snps`) with base quality at least 20 and genotype quality at least 10 (`QUAL > 19 && FMT/GQ >9`) were selected for analysis.
+-->
+
+<!-- ```{r bcf filter, engine = 'bash', eval = FALSE, echo = TRUE} -->
+
+<!-- ## this is one way to filter, i guess but doesn't give nice little report the way that vcf tools does... -->
+
+<!-- # filtering mpileup calls -->
+
+<!-- ### bcftools filter --set-GTs . --include 'QUAL > 19 && FMT/GQ >9' mpileup.call.vcf | \ -->
+
+<!-- ### bcftools view --min-alleles 2 --max-alleles 2 --types snps --apply-filter "PASS" \ -->
+
+<!-- ### --output-type v --output-file mpileup.filtered.vcf -->
+
+<!-- ### # filtering freebayes calls -->
+
+<!-- ### bcftools filter --set-GTs . --include 'QUAL > 19 && FMT/GQ >9' freebayes.call.vcf | \ -->
+
+<!-- ### bcftools view --min-alleles 2 --max-alleles 2 --types snps --apply-filter "PASS" \ -->
+
+<!-- ### --output-type v --output-file freebayes.filtered.vcf -->
+
+<!-- ``` -->

### 2021-03-14 manuscript.Rmd
did not accomplish much

--- 2021-03-06 manuscript.Rmd
@@
 ---
-title: 'Project 3: Variant Calling'
+title: 'Project 3: Variant calling in Burbot (*Lota lota*) using `freebayes` and `mpileup`'
 author: "J Moggridge"
 date: "06/03/2021"
-output: pdf_document
+output: 
+  pdf_document:
+    highlight: kate
+    toc: TRUE
+    toc_depth: 3
+bibliography: references.bib
+link-citations: true
 ---
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+```{r setup, echo=F}
+knitr::opts_chunk$set(warning = F, message = F, eval = T, fig.align = 'center', echo = T)
 ```
 
-## R Markdown
+# Introduction
 
-This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.
+Variant calling is a crucial step in population genetic analysis pipelines.
+Because high-throughput sequencing is inherently noisy, there is a need to apply stringent selection criteria for variants.
+How to decide on these criteria though, is neither simple nor easy to ascertain, as there exists no ground-truth for genomes.
+Indeed, there are many tools for variant calling that differ in their approaches (`bcftools`, `GATK`, `freebayes`, *etc.*).
+These generally have low concordance (cite stuff from hwang), but not in all cases and this may vary depending on the data in question [@hwang2015].
 
-When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
-You can embed an R code chunk like this:
+`bcftools` uses a two step process where `mpileup` first creates the 'pileup' of all reads at a position, and `call` then [@danecek2011, @li2011]
 
-```{r cars}
-summary(cars)
+In this work I compared variant calls made by `bcftools` and `freebayes`.
+Both callers were used to obtain biallelic SNPs and results were compared in terms of the number of SNPs, the overlap between sets, the depth at each position, and their minor allele frequencies.
+
+`freebayes` uses... @garrison2012
+
+------------------------------------------------------------------------
+
+# Methods
+
+<!-- (chmod 755 /scratch/username/directory/ -R). -->
+
+<!-- srun --pty --account="def-nricker" -t 0-03:00:00 --mem=32000 /bin/bash -->
+
+### Dataset
+
+I used data from project \#2 in this work, specifically the reads mapped to the burbot reference with the `bowtie2` aligner.
+I used `bcftools mpileup` and `GATK HaplotypeCaller` to call variants for the ten individuals using these alignments.
+
+I first needed to convert my `.sam` files to `.bam` format and index them, as well as adding read groups using `picard` (necessary for `freebayes`).
+
+```{r sam-to-bam and readgroups, engine = 'bash', eval = FALSE, echo=TRUE}
+#!/bin/bash
+# to convert .sam to .bam, then sort & index
+module load samtools
+for file in bam/*.sam; do
+  name=`echo $file | sed 's/\.sam//'`
+  samtools view -b -S -o $name.bam $file
+  samtools sort $name.bam -o $name.sorted.bam;
+  samtools index $name.sorted.bam
+done
+
+# add readgroups and re-index; show output to verify
+module load picard 
+group=1
+for file in bam/*.sorted.bam
+do
+  label=`echo $file | sed 's/.sorted.bam//'`
+  java -jar $EBROOTPICARD/picard.jar AddOrReplaceReadGroups I=$file \
+    O=$label.rg.sorted.bam RGID=$group RGLB=lib1 RGPL=illumina RGPU=unit1 RGSM=$label
+  samtools index $label.rg.sorted.bam
+  echo "group number: $group; file: $file; read group:"
+  samtools view -H $label.rg.sorted.bam | grep "@RG"
+  group=$((group+1))
+done
 ```
 
-## Including Plots
+### `bcftools mpileup` and `freebayes` variant calling
 
-You can also embed plots, for example:
+I ran `bcftools` variant calling for the 10 .sorted.bam files.
+For `mpileup`, I added arguments to provide annotations for read depth and allelic depth (-a DP,AD), to skip indels, to specify Illumina as the sequencing platform (-P), to consider up to 1000 reads at a locus instead of the default 250 (--max-depth).
+For `call`, I specified to use the multi-allelic caller, to skip indels, and to provide the genotype quality scores (GQ).
 
-```{r pressure, echo=FALSE}
-plot(pressure)
+```{r mpileup, engine = 'bash', eval = FALSE, echo=TRUE}
+module load nixpkgs/16.09  gcc/7.3.0 bcftools/1.9
+# mpileup and variant calling
+bcftools mpileup -Ou -a DP,AD --skip-indels -P ILLUMINA  --max-depth 1000 \
+-f ./burbot_ref/GCA_900302385.1_ASM90030238v1_genomic.fna ./bam/*.rg.sorted.bam | \
+bcftools call -mv --format-fields GQ -o  mpileup.call.vcf
 ```
 
-Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
+I called SNPs with `freebayes` with default settings and annotations for genotype quality. Indels were removed afterwards with `vcffilter`.
+
+```{r freebayes, engine = 'bash', eval = FALSE, echo = TRUE}
+module --force purge; module load nixpkgs/16.09  gcc/7.3.0 freebayes/1.2.0
+# create a list of files to pass to as inputs
+> files.txt
+for file in bam/*.rg.sorted.bam; do echo $file >> files.txt; done
+# free-bayes calling
+freebayes -f burbot_ref/GCA_900302385.1_ASM90030238v1_genomic.fna  \
+  --genotype-qualities --bam-list files.txt | \
+vcffilter -f "TYPE = snp" > freebayes.call.vcf
+```
+
+### Filtering variant calls
+
+Biallelic SNPs (`--min-alleles 2 --max-alleles 2 --remove-indels`) were retained if they had a minimum quality score of 30, minimum genotyping quality score of 10, and been genotyped in at least 50% of individuals (`--minQ 30 --minGQ 10 --max-missing 0.5`).
+
+Given the small number of individuals, I did not feel it necessary to filter by allele frequency.
+
+<!-- (--remove-filtered-all) is like  (--apply-filter "PASS")-->
+
+<!-- (--min-alleles 2 --max-alleles 2) for biallelic -->
+
+<!-- (--minQ 30 --minGQ 10 --max-missing 0.5) for QUAL, GQ, and prop missing data -->
+
+<!-- (-recode --recode-INFO-all) to get a new vcf with annotation data -->
+
+SHOULD ADD FILTER FOR: --maf 0.01 --max-maf 0.99
+
+```{r filtering, engine='bash', eval = FALSE, echo = TRUE}
+module --force purge; module load StdEnv/2020 vcftools/0.1.16
+# filter mpileup calls 
+vcftools --vcf mpileup.call.vcf --minQ 30 --minGQ 10 --max-missing 0.5 \
+  --min-alleles 2 --max-alleles 2 --remove-indels \
+  --remove-filtered-all --recode --recode-INFO-all --out mpileup.q30gq10g5
+# same thing for freebayes
+vcftools --vcf freebayes.call.vcf --minQ 30 --minGQ 10 --max-missing 0.5 \
+  --min-alleles 2 --max-alleles 2 --remove-indels \
+  --remove-filtered-all --recode --recode-INFO-all --out freebayes.q30gq10g5
+```
+
+There was no need to exclude any individuals from the analysis entirely, as none had more that 12% missing data (from `--missing-indv`).
+
+```{r check_missing, engine='bash', eval = FALSE, echo = TRUE}
+# check individuals' missing data
+vcftools --vcf mpileup.q30gq10g5.recode.vcf --missing-indv \
+  --out mpileup.q30gq10g5
+cat mpileup.q30gq10g5.imiss
+vcftools --vcf freebayes.q30gq10g5.recode.vcf --missing-indv \
+  --out freebayes.q30gq10g5
+cat freebayes.q30gq10g5.imiss
+```
+
+### Summarizing SNP calls
+
+I counted and compared the biallelic SNP calls from each caller with `bcftools stats` and `isec`.
+
+```{r count_snps, engine = 'bash', eval = FALSE, echo = TRUE}
+module load StdEnv/2020 bcftools/1.10.2
+# count the snps: 40,265 and 29,176
+bcftools stats mpileup.q30gq10g5.recode.vcf | grep "number of SNPs:" | cut -f4
+bcftools stats freebayes.q30gq10g5.recode.vcf | grep "number of SNPs:" | cut -f4
+
+# prepare mpileup snps by zipping and indexing
+bgzip mpileup.q30gq10g5.recode.vcf
+bcftools index mpileup.q30gq10g5.recode.vcf.gz
+# Prepare freebayes snps by zipping and indexing
+bgzip freebayes.q30gq10g5.recode.vcf
+bcftools index freebayes.q30gq10g5.recode.vcf.gz
+
+# Compute intersection of 2 files with isec
+bcftools isec -p isec mpileup.q30gq10g5.recode.vcf.gz freebayes.q30gq10g5.recode.vcf.gz
+
+# Get overlap data from isec into tsv format
+> Overlap.vals
+for file in isec/*.vcf
+do 
+  bcftools stats $file| grep "number of SNPs:" | cut -f4 >> Overlap.vals
+done
+echo $'unique to mpileup\nunique to freebayes\nshared\nshared' > Overlap.names
+paste  Overlap.names Overlap.vals > Overlap.tsv
+sed -i '1s/^/SNPs\tCount\n/' Overlap.tsv
+```
+
+Using `vcftools`:
+
+-   Allele frequency at each loci across individuals
+
+```{r vcf_freq, engine = 'bash', eval = FALSE, echo = TRUE}
+vcftools --gzvcf mpileup.q30gq10g5.recode.vcf.gz --freq \
+  --out mpileup.q30gq10g5
+vcftools --gzvcf freebayes.q30gq10g5.recode.vcf.gz --freq \
+  --out freebayes.q30gq10g5
+```
+
+-   Sum depth at each locus across individuals
+
+```{r vcf_site_depth, engine = 'bash', eval = FALSE, echo = TRUE}
+vcftools --gzvcf mpileup.q30gq10g5.recode.vcf.gz --site-depth \
+  --out mpileup
+vcftools --gzvcf freebayes.q30gq10g5.recode.vcf.gz --site-depth \
+  --out freebayes
+```
+
+-   Mean depth for each individual across all loci, relatedness, and genotypes
+
+```{r vcf_indv_depth, engine = 'bash', eval = FALSE, echo = TRUE}
+# mean depth per individual
+vcftools --gzvcf mpileup.q30gq10g5.recode.vcf.gz --depth --out mpileup
+vcftools --gzvcf freebayes.q30gq10g5.recode.vcf.gz --depth --out freebayes
+
+## Relatedness
+vcftools --gzvcf mpileup.q30gq10g5.recode.vcf.gz --relatedness --out mpileup
+vcftools --gzvcf freebayes.q30gq10g5.recode.vcf.gz --relatedness --out freebayes
+
+## Convert to (0,1,2) format
+vcftools --gzvcf mpileup.q30gq10g5.recode.vcf.gz --012 --out mpileup
+vcftools --gzvcf freebayes.q30gq10g5.recode.vcf.gz --012 --out freebayes
+```
+
+<!-- quality, as shown in class, using what you think is a reasonable threshold.
+Compare the resulting vcf files in several ways.
+First, *compare the number of SNPs* and the *overlap of these SNPs across files*.
+-->
+
+<!-- Then, plot the *minor allele frequency* and depth of all SNPs (summed across individuals), and discuss the variation across loci. Compare across vcfs produced by different variant callers. -->
+
+    # get data down from cedar in local terminal:
+    scp jmoggrid@cedar.computecanada.ca:/scratch/jmoggrid/Project3/Overlap.tsv .
+    scp jmoggrid@cedar.computecanada.ca:/scratch/jmoggrid/Project3/*.frq .
+    scp jmoggrid@cedar.computecanada.ca:/scratch/jmoggrid/Project3/*.idepth .
+    scp jmoggrid@cedar.computecanada.ca:/scratch/jmoggrid/Project3/*.ldepth .
+
+------------------------------------------------------------------------
+
+### Statistics
+
+I created figures and summary tables in the `R` language
+
+```{r load data, message=FALSE}
+#! (R code)
+library(tidyverse, quietly = T)
+library(ggbeeswarm)
+library(patchwork)
+library(rcartocolor)
+library(gridExtra)
+library(grid)
+
+# Overlap counts
+Overlap <- read_delim('./data/Overlap.tsv', '\t') %>%  
+  distinct() %>% 
+  bind_rows(
+    tibble(SNPs = c('mpileup', 'freebayes'), 
+           Count = c(.[1,2][[1]] + .[3,2][[1]], 
+                     .[2,2][[1]] + .[3,2][[1]]))
+    ) %>%  arrange(desc(Count))
+
+# Allele frequencies
+cnames <- c('chrom','pos', 'n_alleles','n_chr','a1','a2')
+AlleleFreq <-  
+  bind_rows(
+    read_tsv('./data/freebayes.q30gq10g5.frq', col_names = cnames, skip = 1L) %>% 
+      mutate(caller = 'freebayes'),
+    read_tsv('./data/mpileup.q30gq10g5.frq', col_names = cnames, skip = 1L) %>% 
+      mutate(caller = 'mpileup')
+  ) %>% 
+  mutate(allele = str_remove(a1, ':.+'), frequency = str_remove(a1, '.:'),
+         allele2 = str_remove(a2, ':.+'), frequency2 = str_remove(a2, '.:')
+  ) %>% 
+  mutate(across(contains('freq'), as.numeric))
+
+# Depth
+cnames <- c('individual', 'n sites', 'mean_depth')
+IndvDepth <- bind_rows(
+    read_tsv('./data/mpileup.idepth', col_names = cnames, skip = 1L) %>%
+      mutate(caller = 'mpileup'),
+    read_tsv('./data/freebayes.idepth', col_names = cnames, skip = 1L) %>%
+      mutate(caller = 'freebayes')
+  ) %>% 
+  mutate(individual = str_remove_all(individual, 'bam.|.bowtie.burbot'),
+         caller = as_factor(caller)) %>% 
+  arrange(individual)
+
+# Site Depth
+cnames <- c("chr", "pos", "allele_depth", "var_depth")
+SiteDepth <- bind_rows(
+  read_tsv('./data/mpileup.ldepth', col_names = cnames, skip = 1L) %>%
+    mutate(caller = 'mpileup'),
+  read_tsv('./data/freebayes.ldepth', col_names = cnames, skip = 1L) %>%
+    mutate(caller = 'freebayes')
+) %>% mutate(caller = as_factor(caller))
+
+```
+
+```{r figs, echo=F}
+# Allele frequency histogram (alleles stacked)
+fig_allelefreq <- 
+  ggplot(AlleleFreq %>% filter(frequency != 0), 
+         aes(x = frequency*10, fill=allele)) +
+  geom_histogram(bins = 20, colour = 'white',) +
+  facet_wrap(~caller) +
+  rcartocolor::scale_fill_carto_d() +
+  labs(x = 'minor allele frequency', y='loci', fill = 'Allele') +
+  theme_bw()
+# All allele freqs (including maf = 0, boo)
+fig_allelefreq_all <- 
+  ggplot(AlleleFreq, aes(x = G1.freq)) +
+  geom_histogram(bins = 20, colour = 'white',) +
+  facet_wrap(~caller) +
+  scale_fill_carto_d() +
+  labs(x = 'minor allele frequency', y='loci') +
+  theme_bw()
+# The distribution of depth per-site summed across individuals
+fig_sitedepth <- 
+  ggplot(SiteDepth, aes(x = allele_depth, fill = caller, color = caller)) +
+  geom_density(alpha = 0.12) +
+  geom_vline(aes(xintercept = mean(allele_depth), color = caller), lty = 2) +
+  scale_x_log10() +
+  labs(x = 'depth per locus', y='density') +
+  guides(colour = guide_legend(override.aes = list(size=1, pch=15))) +
+  theme_bw()
+# The distribution of mean depth per site within individuals
+set.seed(45)
+fig_indvdepth <-
+  ggplot(IndvDepth, aes(y = mean_depth, x = caller)) +
+  geom_violin(alpha = 0.1, draw_quantiles = c(.25, .5, .75), trim = F) +
+  geom_quasirandom(aes(color = individual), size = 3, width = 0.3) +
+  scale_color_carto_d(palette = 2) +
+  guides(colour = guide_legend(override.aes = list(size = 4, pch = 16))) +
+  facet_wrap( ~ caller, scales = 'free_x') +
+  labs(y = 'mean depth per locus', x = '', color = 'Individual') +
+  theme_bw() +
+  theme(
+    axis.text.x = element_blank(),
+    axis.ticks.x = element_blank(),
+    legend.position = 'right'
+  )
+```
+
+------------------------------------------------------------------------
+
+\newpage
+
+# Results
+
+Biallelic SNPs identified by each variant caller were counted and compared (table 1).
+`mpileup` called `r scales::comma(Overlap[1,2][[1]])` SNPs, `r round(Overlap[1,2][[1]] * 100 / Overlap[2,2][[1]], 2) - 100` % more than the `r scales::comma(Overlap[2,2][[1]])` called by `freebayes`.
+Of these, `r scales::comma(Overlap[3,2][[1]])` were shared between both sets, representing `r round(Overlap[3,2][[1]] * 100 /Overlap[1,2][[1]], 2)` % of `mpileup` calls and `r round(Overlap[3,2][[1]] * 100 /Overlap[2,2][[1]], 2)` % of `freebayes` calls.
+Additionally, `r scales::comma(Overlap[4,2][[1]])` SNPs were unique to `mpileup` and `r scales::comma(Overlap[5,2][[1]])` to `freebayes`.
+
+The allele frequency....
+
+The sum depth per locus across all individuals is generally in the range \~ 50-200 reads, with both callers showing similar distributions (fig 1.C).
+The mean depth for `mpileup` was 173.7 reads and for `freebayes` 162.2 reads.
+Both callers have some outliers in terms of sum depth across individuals: the maximum depth for `freebayes` was 56,564 reads, while for `freebayes` it was 10,046 reads (as the max-depth consideration of 1000 was applied here).
+
+------------------------------------------------------------------------
+
+```{r fig1, echo=F, fig.align = 'center', fig.width=9, fig.height=5.5, fig.cap="(A) Distribution of depth per locus summed across all individuals (note log-scaled x-axis).
+(B) Mean depth per locus for each individual.
+(C) Distribution of minor allele frequency; SNPs with frequency = 0 or 1 were omitted.
+(D)Total, shared, and unique SNPs called by mpileup and freebayes"}
+(
+  (fig_allelefreq + fig_indvdepth) /
+  (fig_sitedepth + gridExtra::tableGrob(Overlap, rows = NULL))
+) &
+  plot_layout(guides = 'keep') &
+  plot_annotation(tag_levels = 'A')
+```
+
+\newpage
+
+# Discussion
+
+------------------------------------------------------------------------
+
+\newpage
+
+# References

### 2021-03-18 manuscript.Rmd
need to find the stuff i deleted

--- 2021-03-06 manuscript.Rmd
@@
 ---
-title: 'Project 3: Variant Calling'
+title: 'Project 3: Variant calling in Burbot (*Lota lota*) using `freebayes` and `mpileup`'
 author: "J Moggridge"
 date: "06/03/2021"
-output: pdf_document
+output: 
+  pdf_document:
+    highlight: kate
+bibliography: references.bib
+link-citations: true
 ---
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+```{r setup, echo=F}
+knitr::opts_chunk$set(warning = F, message = F)
 ```
 
-## R Markdown
+------------------------------------------------------------------------
 
-This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.
+# Introduction
 
-When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
-You can embed an R code chunk like this:
+<!-- purpose of variant calling -->
 
-```{r cars}
-summary(cars)
+The analysis of genetic variants (single-nucleotide polymorphisms -SNPs and insertions/deletions - indels), by high-throughput sequencing has applications in studies of phylogeny and population structure, quantitative-trait loci, genome-wide associations, among others.
+Such polymorphic sites are also of practical interest in medical genomics, *eg*.
+to discover disease-causing mutations [@cornish2015], and in agriculture, where SNPs can be used as genetic markers to guide breeding of crops [@yao2020] and livestock [@ni2015].
+As the cost of sequencing has decreased greatly and computational resources have improved, these techniques have become increasingly popular.
+
+The identification of polymorphic sites from sequencing data, known as variant calling, is performed after alignment of reads to a reference genome.
+Because high-throughput sequencing and read-mapping are error-prone, it is challenging but essential to differentiate true variants from artefacts, particularly in low-coverage data [@nielsen2011].
+Many tools for variant calling have been created for this classification task (`bcftools`, `GATK`, `freebayes`, *etc.*), though these differ in their statistical models for inferring genotypes.
+Further filtering of variants is generally applied after calling, and how this is performed will vary depending on the nature of the study.
+
+In the current work I compared variant calling by two tools: `bcftools mpileup/call` (previously `samtools`) and `freebayes`.
+Both `bcftools` and `freebayes` apply Bayesian models to estimate genotype posterior probabilities, though they differ in how this is implemented.
+`bcftools` uses a two step process where `mpileup` first creates the 'pileup' of all reads at variant sites, and then `call` is used to infer genotypes at these [@li2011c; @danecek2021a].
+This implementation has the advantage of being very fast.
+`freebayes` a uses different approach to inferring haplotypes in that it does not use the alignment but the full reads [@garrison2012a; @howtofa].
+
+In this work, `bcftools` and `freebayes` variant callers were used to obtain biallelic SNPs and results were compared in terms of the number of SNPs, the overlap between sets, the depth at each position, and their minor allele frequencies.
+
+Best method could depend on nature of the sequencing data: eg. whole-genome, RAD-seq, RNA-seq, scRNA-seq, etc, and the depth of coverage.
+
+------------------------------------------------------------------------
+
+# Methods
+
+Data can be found on `cedar` in `/scratch/jmoggrid/Project3/`. <!-- (chmod 755 /scratch/username/directory/ -R). --> <!-- srun --pty --account="def-nricker" -t 0-03:00:00 --mem=32000 /bin/bash -->
+
+### Dataset
+
+I used data from project \#2 in this work, specifically the reads mapped to the burbot reference with the `bowtie2` aligner.
+I used `bcftools mpileup` and `GATK HaplotypeCaller` to call variants for the ten individuals using these alignments.
+
+I first needed to convert my `.sam` files to `.bam` format and index them, as well as adding read groups using `picard` (necessary for `freebayes`).
+
+```{r sam-to-bam and readgroups, engine = 'bash', eval = FALSE, echo=TRUE}
+#!/bin/bash
+# to convert .sam to .bam, then sort & index
+module load samtools
+for file in bam/*.sam; do
+  name=`echo $file | sed 's/\.sam//'`
+  samtools view -b -S -o $name.bam $file
+  samtools sort $name.bam -o $name.sorted.bam;
+  samtools index $name.sorted.bam
+done
+
+# add readgroups and re-index; show output to verify
+module load picard 
+group=1
+for file in bam/*.sorted.bam
+do
+  label=`echo $file | sed 's/.sorted.bam//'`
+  java -jar $EBROOTPICARD/picard.jar AddOrReplaceReadGroups I=$file \
+    O=$label.rg.sorted.bam RGID=$group RGLB=lib1 RGPL=illumina RGPU=unit1 RGSM=$label
+  samtools index $label.rg.sorted.bam
+  echo "group number: $group; file: $file; read group:"
+  samtools view -H $label.rg.sorted.bam | grep "@RG"
+  group=$((group+1))
+done
 ```
 
-## Including Plots
+### `bcftools mpileup` and `freebayes` variant calling
 
-You can also embed plots, for example:
+I ran `bcftools` variant calling for the 10 .sorted.bam files.
+For `mpileup`, I added arguments to provide annotations for read depth and allelic depth (-a DP,AD) and to specify Illumina as the sequencing platform (-P).
+For `call`, I provided arguments to specify the use of the multi-allelic caller, 'variants only', to skip indels, and to provide the genotype quality scores (GQ).
 
-```{r pressure, echo=FALSE}
-plot(pressure)
+```{r mpileup, engine = 'bash', eval = FALSE, echo=TRUE}
+module load nixpkgs/16.09  gcc/7.3.0 bcftools/1.9
+# mpileup and variant calling
+bcftools mpileup -Ou -a DP,AD -P ILLUMINA \
+-f ./burbot_ref/GCA_900302385.1_ASM90030238v1_genomic.fna ./bam/*.rg.sorted.bam | \
+bcftools call -mv --format-fields GQ --skip-variants indels -o  mpileup.call.vcf
+bcftools stats mpileup.call.vcf > mpileup.call.stats
 ```
 
-Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
+I called SNPs using `freebayes` with default settings; annotations for genotype quality were requested with `--genotype-qualities`. Output was limited to SNPs by piping to `vcffilter`.
+
+```{r freebayes, engine = 'bash', eval = FALSE, echo = TRUE}
+module load freebayes/1.2.0
+# create a list of 10 filenames to pass to as input to freebayes
+> files.txt
+for file in bam/*.rg.sorted.bam; do echo $file >> files.txt; done
+# free-bayes calling for 10 files and restricting vcf output to SNPs w vcffilter
+freebayes -f burbot_ref/GCA_900302385.1_ASM90030238v1_genomic.fna  \
+  --genotype-qualities --bam-list files.txt | \
+vcffilter -f "TYPE = snp" > freebayes.call.vcf
+bcftools stats freebayes.call.vcf > freebayes.call.vcf
+```
+
+### Filtering variant calls
+
+I used the `vcftools` library to filter variant calls.
+Biallelic SNPs (`--min-alleles 2 --max-alleles 2 --remove-indels`) were retained if they had a quality score of at least 20 (`--minQ 30`).
+Genotypes were retained at these sites if they had genotyping quality score of at least 10 (`--minGQ 10`).
+I found the vast majority of variants to be shared among all 10 individuals, so I applied filters for the maximum and minimum minor allele frequencies, such that these would not be either zero or one (`–maf 0.01 –max-maf 0.99`).
+Any calls that did not receive a 'PASS' flag were discarded (`--remove-filtered-all`).
+
+```{r filtering, engine='bash', eval = FALSE, echo = TRUE}
+module --force purge; module load StdEnv/2020 vcftools/0.1.16
+
+# filter mpileup calls 
+vcftools --vcf mpileup.call.vcf --minQ 20 --minGQ 10 \
+  --min-alleles 2 --max-alleles 2 --maf 0.01 --max-maf 0.99 \
+  --remove-filtered-all --recode --recode-INFO-all \
+  --out mpileup.filter
+
+# same thing for freebayes
+vcftools --vcf freebayes.call.vcf --minQ 20 --minGQ 10 \
+  --min-alleles 2 --max-alleles 2 --maf 0.01 --max-maf 0.99 \
+  --remove-filtered-all --recode --recode-INFO-all \
+  --out freebayes.filter
+```
+
+There was no need to exclude any individuals from the analysis, as none had more that 22.5% missing data in the filtered `mpileup` calls, while the proportions of missing data in `freebayes` calls was roughly half as large (from `--missing-indv`).
+
+```{r check_missing, engine='bash', eval = FALSE, echo = TRUE}
+# check individuals' missing data
+vcftools --vcf mpileup.filter.recode.vcf --missing-indv \
+  --out mpileup.filter
+vcftools --vcf freebayes.filter.recode.vcf --missing-indv \
+  --out freebayes.filter
+```
+
+### Summarizing SNP calls
+
+I counted the biallelic SNP calls from each caller with `bcftools stats`.
+
+```{r count_snps, engine = 'bash', eval = FALSE, echo = TRUE}
+module load StdEnv/2020 bcftools/1.10.2
+bcftools stats mpileup.filter.recode.vcf | grep "number of SNPs:" | cut -f4
+bcftools stats freebayes.filter.recode.vcf | grep "number of SNPs:" | cut -f4
+```
+
+The overlap between sets of SNPs was computed with `bcftools isec`.
+
+```{r overlap_in_sets, engine = 'bash', eval = FALSE, echo = TRUE}
+# prepare files for isec by zipping and re-indexing
+bgzip mpileup.filter.recode.vcf
+bcftools index mpileup.filter.recode.vcf.gz
+bgzip freebayes.filter.recode.vcf
+bcftools index freebayes.filter.recode.vcf.gz
+# Compute intersection of 2 files with isec
+bcftools isec -p isec mpileup.filter.recode.vcf.gz freebayes.filter.recode.vcf.gz
+```
+
+<!-- # Get overlap data from isec into tsv format -->
+
+<!-- > Overlap.vals -->
+
+<!-- for file in isec/*.vcf -->
+
+<!-- do  -->
+
+<!--   bcftools stats $file| grep "number of SNPs:" | cut -f4 >> Overlap.vals -->
+
+<!-- done -->
+
+<!-- echo $'unique to mpileup\nunique to freebayes\nshared\nshared' > Overlap.names -->
+
+<!-- paste  Overlap.names Overlap.vals > Overlap.tsv -->
+
+<!-- sed -i '1s/^/SNPs\tCount\n/' Overlap.tsv -->
+
+Using `vcftools`, I computed the following:
+
+-   Allele frequency at each loci across individuals
+
+```{r vcf_freq, engine = 'bash', eval = FALSE, echo = TRUE}
+vcftools --gzvcf mpileup.filter.recode.vcf.gz --freq \
+  --out mpileup.filter
+vcftools --gzvcf freebayes.filter.recode.vcf.gz --freq --out freebayes.filter
+```
+
+-   Sum depth at each locus across individuals
+
+```{r vcf_site_depth, engine = 'bash', eval = FALSE, echo = TRUE}
+vcftools --gzvcf mpileup.filter.recode.vcf.gz --site-depth --out mpileup
+vcftools --gzvcf freebayes.filter.recode.vcf.gz --site-depth --out freebayes
+```
+
+```{=html}
+<!-- # get data down from cedar in local terminal:
+ scp jmoggrid@cedar.computecanada.ca:/scratch/jmoggrid/Project3/Overlap.tsv .
+ scp jmoggrid@cedar.computecanada.ca:/scratch/jmoggrid/Project3/*.frq . 
+ scp jmoggrid@cedar.computecanada.ca:/scratch/jmoggrid/Project3/*.idepth . 
+ scp jmoggrid@cedar.computecanada.ca:/scratch/jmoggrid/Project3/*.ldepth .  -->
+```
+
+------------------------------------------------------------------------
+
+### Statistics
+
+I created figures and summary tables from the data generated above using the `R` language.
+
+```{r load_data, message=FALSE}
+#! {R code}
+library(tidyverse, quietly = T)
+
+# Overlap counts table; add
+Overlap <- read_delim('./data/Overlap.tsv', '\t') %>%  
+  distinct() %>% 
+  bind_rows(tibble(SNPs = c('mpileup', 'freebayes'), 
+                   Count = c(.[1,2][[1]] + .[3,2][[1]], 
+                             .[2,2][[1]] + .[3,2][[1]]))) %>%  
+  arrange(desc(Count))
+```
+
+```{r load_data2, message=FALSE, echo=FALSE}
+# Allele frequencies
+cnames <- c('chrom','pos', 'n_alleles','n_chr','a1','a2')
+AlleleFreq <-  
+  bind_rows(
+    read_tsv('./data/mpileup.filter.frq', col_names = cnames, skip = 1L) %>% 
+      mutate(caller = 'mpileup'),
+    read_tsv('./data/freebayes.filter.frq', col_names = cnames, skip = 1L) %>% 
+      mutate(caller = 'freebayes')
+  ) %>% 
+  mutate(allele = str_extract(a1, '[A-Z]+'), 
+         frequency = str_remove_all(a1, '[A-Z]+:'),
+         allele2 = str_extract(a2, '[A-Z]+'), 
+         frequency2 = str_remove_all(a2, '[A-Z]+:')
+  ) %>% 
+  mutate(across(contains('freq'), as.numeric)) %>%
+  mutate(
+    maf = ifelse(frequency > 0.5, 1 - frequency, frequency),
+    minor_allele = ifelse(maf == frequency, allele, allele2)
+  )
+
+# Depth
+cnames <- c('individual', 'n sites', 'mean_depth')
+IndvDepth <- bind_rows(
+    read_tsv('./data/mpileup.idepth', col_names = cnames, skip = 1L) %>%
+      mutate(caller = 'mpileup'),
+    read_tsv('./data/freebayes.idepth', col_names = cnames, skip = 1L) %>%
+      mutate(caller = 'freebayes')
+  ) %>% 
+  mutate(individual = str_remove_all(individual, 'bam.|.bowtie.burbot'),
+         caller = as_factor(caller)) %>% 
+  arrange(individual)
+
+# Site Depth
+cnames <- c("chr", "pos", "allele_depth", "var_depth")
+SiteDepth <- bind_rows(
+  read_tsv('./data/mpileup.ldepth', col_names = cnames, skip = 1L) %>%
+    mutate(caller = 'mpileup'),
+  read_tsv('./data/freebayes.ldepth', col_names = cnames, skip = 1L) %>%
+    mutate(caller = 'freebayes')
+) %>% mutate(caller = as_factor(caller))
+
+vcf_df <- SiteDepth %>%
+  rename(chrom = chr) %>% 
+  left_join(AlleleFreq, by = c('chrom', 'pos', 'caller'))
+
+cnames <- c('id', 'quality', 'n_snps', 'ts', 'tv', 'indels')
+qual_df <- bind_rows(
+  read_tsv("./data/mpileup.quality.stats",
+                 skip = 1L, col_names = cnames) %>% 
+  mutate(caller = "mpileup"),
+  read_tsv("./data/freebayes.quality.stats",
+                 skip = 1L, col_names = cnames) %>% 
+  mutate(caller = "freebayes")
+  ) %>% 
+  mutate(caller = as_factor(caller))
+
+```
+
+```{r figs, echo=F}
+## this chunk contains code for all the plots
+library(ggbeeswarm)
+library(patchwork)
+library(rcartocolor)
+library(gridExtra)
+library(grid)
+
+# Allele frequency histogram (alleles stacked)
+fig_allelefreq <- 
+  ggplot(AlleleFreq) +
+  geom_histogram(aes(x = maf), bins = 20, alpha = 0.8) +
+  facet_wrap(caller~., scales = 'free_y') +
+  labs(x = 'minor allele frequency', y='loci') +
+  theme_bw()
+
+# The distribution of depth per-site summed across individuals
+fig_sitedepth <- 
+  ggplot(SiteDepth, aes(x = allele_depth, fill = caller, color = caller)) +
+  geom_density(alpha = 0.2) +
+  geom_vline(aes(xintercept = mean(allele_depth), color = caller),
+             lty = 2) +
+  scale_x_log10() +
+  labs(x = 'depth per locus\n(sum of 10 individuals)', y = 'density') +
+  scale_color_carto_d(palette = 2, direction = -1) +
+  scale_fill_carto_d(palette = 2, direction = -1) +
+  guides(colour = guide_legend(override.aes = list(size=1, pch=15))) +
+  theme_bw() +
+  theme(legend.position = c(.8,.7))
+
+# The distribution of mean depth per site within individuals
+set.seed(45)
+fig_indvdepth <-
+  ggplot(IndvDepth, aes(y = mean_depth, x = caller)) +
+  geom_violin(alpha = 0.1, draw_quantiles = c(.25, .5, .75), trim = F) +
+  geom_quasirandom(aes(color = individual), size = 3, width = 0.3) +
+  scale_color_carto_d(palette = 2) +
+  guides(colour = guide_legend(override.aes = list(size = 4, pch = 16))) +
+  facet_wrap( ~ caller, scales = 'free_x') +
+  labs(y = 'mean depth per locus', x = '', color = 'Individual') +
+  theme_bw() +
+  theme(
+    axis.text.x = element_blank(),
+    axis.ticks.x = element_blank(),
+    legend.position = 'right'
+  )
+
+# Allele depth over maf, coloured by n chromosomes
+fig_alleledepth_over_maf <- 
+  ggplot(vcf_df, aes(x=maf, y=allele_depth, color = n_chr)) +
+  geom_point(size = 0.5, alpha = 0.7) +
+  scale_y_log10() +
+  facet_wrap(~caller) +
+  scale_color_viridis_c() +
+  labs(x = "Minor allele frequency",
+       y = "Allele depth", 
+       color="Number of\nchromosomes") +
+  theme_bw()
+
+# Quality score distribution by caller fig
+fig_quality_dist <- qual_df %>% 
+  mutate(QUAL = round(quality/25)*25) %>% 
+  group_by(QUAL, caller) %>% 
+  summarise(n_snps = sum(n_snps),
+            ts  = sum(ts),
+            tv = sum(tv)) %>% 
+  pivot_longer(cols = c(ts,tv)) %>% 
+  ggplot(aes(x = QUAL, y = value, fill = name)) +
+  geom_col(width = 19) +
+  facet_wrap(~caller) +
+  scale_fill_carto_d(palette = 2) +
+  labs(x="quality score", y = "loci", fill = NULL) +
+  scale_y_log10() +
+  theme_bw()
+
+```
+
+------------------------------------------------------------------------
+
+\newpage
+
+# Results
+
+From the ten burbots in our sample, `mpileup` called a total of 132,046 SNPs, 173 of which were multiallelic, and with a transition:transversion ratio (ts:tv) of 0.81; `freebayes`called a total of 106,973 SNPs, with 1761 being multiallelic, and ts:tv of 0.44 (table 1).
+After applying filters for these were reduced to 11,115 SNPs with a 0.87 transition:transversion ratio for `mpileup` calls and 6,722 SNPs with 1.09 transition:transversion ratio for `freebayes` calls.
+The sets of biallelic SNPs identified by each caller were compared (table 2): `mpileup` called `r round(Overlap[1,2][[1]] * 100 / Overlap[2,2][[1]], 2) - 100` % more biallelic SNPs than `freebayes`.
+Of these, a large proportion (`r scales::comma(Overlap[3,2][[1]])`) were shared between both sets, representing `r round(Overlap[3,2][[1]] * 100 /Overlap[1,2][[1]], 2)` % of `mpileup` calls and `r round(Overlap[3,2][[1]] * 100 /Overlap[2,2][[1]], 2)` % of `freebayes` calls.
+Additionally, `r Overlap[4,2][[1]]` SNPs were unique to `mpileup` and `r Overlap[5,2][[1]]` to `freebayes`.
+
+The distribution minor allele frequencies of SNPs from `mpileup` shows a U-shape with peaks at 0.05 and 0.5, whereas those from `freebayes` peak at 0.05 and are otherwise evenly distributed (fig.1A).
+The sum depth per locus across all individuals is generally in the range of 50-200 reads, with both callers showing similar distributions (fig.1B).
+The mean depth for `mpileup` was 173.7 reads and for `freebayes` 162.2 reads.
+Both callers have some extreme outliers in terms of site depth: the maximum depth for `freebayes` was 16,162 reads, while for `mpileup` it was only 2522 reads.
+The minimum site depth was 5 and 3 for `freebayes` and `mpileup` respectively and `mpileup` has a greater proportion of low-depth sites.
+
+------------------------------------------------------------------------
+
+```{r table1_filtering, echo=FALSE}
+tribble(
+  ~ Caller, ~ `raw SNPs`, ~ `raw ts/tv`, ~ `post-filter SNPs`, ~ `post-filter ts/tv`,
+  "mpileup", 132046, 0.81, 11115, 0.87,
+  "freebayes", 106973, 0.44, 6722, 1.09
+) %>% pander::pander(caption = "Raw and filtered SNP counts and transition/transversion ratios for calls by mpileup and freebayes.")
+```
+
+```{r table2_overlap, echo=F}
+Overlap %>% pander::pander(caption = "Total, shared, and unique SNPs called by mpileup and freebayes (after filtering).")
+```
+
+```{r fig1, echo=F, fig.align = 'center', fig.width=8, fig.height=3.5, fig.cap="(A) Distribution of minor allele frequency; SNPs with frequency = 0 or 1 were omitted.
+(B) Distribution of depth per locus summed across all individuals (note log-scaled x-axis)."}
+
+(fig_allelefreq + fig_sitedepth) &
+plot_layout(guides = 'keep') &
+plot_annotation(tag_levels = 'A')
+```
+
+\newpage
+
+# Discussion
+
+<!-- Sites with low quality were discarded -->
+
+<!-- Kept all allele freq, even rare ones ... rare ones are cool because reference... depends of system of interest though ... -->
+
+<!-- Sites with extremely large depth could represent paralogous regions? -->
+
+Comparison of callers: Hwang2015; Cornish2015
+
+-   Reflect on results
+
+-   Concurrence with other findings
+
+    -   If not, why not (data...)
+
+-   Characteristics of the data affect variant calls?
+
+    -   Reference was highly fragmented, though \~85 % reads mapped to the reference in all individuals (in earlier work)
+
+    -
+
+-   Other features of the VCF that are notable ?
+The mean read depth across all sites is larger for all individuals in the `freebayes` SNPs (fig.
+2).
+Perhaps this is because freebayes is less tolerant of low-coverage sites in it's estimation of genotype likelihood/ quality / ....
+
+-   What other filters:
+
+    -   Depth.... high coverage can indicate a paralogous region (duplicated)
+
+    -   Might be interested in triallelic snps or indels, depending on the study...
+
+    -   If looking for rare variants we would perhaps want to not use filters and only look at positions that are relevant to the process or disease of interest [@cornish2015].
+
+Both `freebayes` and `mpileup` apply Bayes' theorem in their algorithms for SNP calling.
+
+mpileup considers the bases at a variant position
+
+freebayes uses a window approach that differs from other bayesian callers which treat the alignment similarly...
+
+cornish
+
+-   callers have high positive predictive value (TP/TP+FP) but low sensitivity (TP/P).
+
+-   Freebayes identified more raw SNPs (far more false positive) in the human exome data; but mpileup and freebayes had similar counts after filtering.
+
+-   Freebayes did not perform as well as mpileup.
+
+------------------------------------------------------------------------
+
+# Supplemental Figures
+
+```{r fig_allele_maf_nchr, echo=F, fig.align='center', fig.height=4, fig.cap="Allele depth and minor allele frequency do not show any correlation; minor allele frequency is constrained to specific values by the number of chromosomes represented"}
+fig_alleledepth_over_maf
+```
+
+## Mean depth by subject
+
+Mean depth for each individual across all loci was computed as follows:
+
+```{r vcf_indv_depth, engine = 'bash', eval = FALSE, echo = TRUE}
+vcftools --gzvcf mpileup.filter.recode.vcf.gz --depth --out mpileup
+vcftools --gzvcf freebayes.filter.recode.vcf.gz --depth --out freebayes
+```
+
+We can see the all individuals have greater mean depth at SNPs called by `freebayes` than `bcftools mpileup` (fig. 4)
+
+```{r fig_indv_depth, echo=F, fig.align = 'center', fig.width=5, fig.height=2.9, fig.cap="Mean depth per locus for each individual is greater for SNPs called by  freebayes than bcftools."}
+fig_indvdepth
+```
+
+### Quality scores
+
+I obtained quality scores for each loci from bctools stats
+
+```{r engine = 'bash', eval=FALSE}
+## quality scores...
+bcftools stats mpileup.filter.recode.vcf > mpileup.filter.recode.stats
+grep -E "Quality" mpileup.filter.recode.stats | cut -f1 --complement > mpileup.quality.stats
+grep ^QUAL mpileup.filter.recode.stats | cut -f1 --complement >> mpileup.quality.stats bcftools stats freebayes.filter.recode.vcf > freebayes.filter.recode.stats
+grep -E "Quality" freebayes.filter.recode.stats | cut -f1 --complement > freebayes.quality.stats
+grep ^QUAL freebayes.filter.recode.stats | cut -f1 --complement >> freebayes.quality.stats
+```
+
+```{r fig3, echo=F, fig.align = 'center', fig.width=5, fig.height=2.9, fig.cap="Minor allele fre"}
+fig_alleledepth_over_maf
+```
+
+```{r fig4, echo=F, fig.height = 3, fig.align = 'center', fig.width=5, fig.height=2.9, fig.cap="The quality score distributions of freebayes and bcftools mpileup "}
+fig_quality_dist
+```
+
+<!-- ## Relatedness -->
+
+<!-- vcftools --gzvcf mpileup.filter.recode.vcf.gz --relatedness --out mpileup -->
+
+<!-- vcftools --gzvcf freebayes.filter.recode.vcf.gz --relatedness --out freebayes -->
+
+<!-- ## Convert to (0,1,2) format -->
+
+<!-- vcftools --gzvcf mpileup.filter.recode.vcf.gz --012 --out mpileup -->
+
+<!-- vcftools --gzvcf freebayes.filter.recode.vcf.gz --012 --out freebayes -->
+
+<!-- quality, as shown in class, using what you think is a reasonable threshold.
+Compare the resulting vcf files in several ways.
+First, *compare the number of SNPs* and the *overlap of these SNPs across files*.
+-->
+
+<!-- Then, plot the *minor allele frequency* and depth of all SNPs (summed across individuals), and discuss the variation across loci. Compare across vcfs produced by different variant callers. -->
+
+\newpage
+
+# References
+
+```{=html}
+<!-- Zeggini, E., & Morris, A. (2015). Assessing Rare Variation in Complex Traits Design and Analysis of Genetic Studies (1st ed. 2015.). Springer 
+New York. <https://doi.org/10.1007/978-1-4939-2824-8> -->
+```

### 2021-03-18 manuscript.Rmd
fixed missing text

@@
 Both `bcftools` and `freebayes` apply Bayesian models to estimate genotype posterior probabilities, though they differ in how this is implemented.
-`bcftools` uses a two step process where `mpileup` first creates the 'pileup' of all reads at variant sites, and then `call` is used to infer genotypes at these [@li2011c; @danecek2021a].
+`bcftools` uses a two step process where `mpileup` first creates the 'pileup' of all reads at variant sites, and then `call` is used to infer genotypes at these [@li2011, @danecek2021].
 This implementation has the advantage of being very fast.
-`freebayes` a uses different approach to inferring haplotypes in that it does not use the alignment but the full reads [@garrison2012a; @howtofa].
+`freebayes` a uses different approach to inferring haplotypes in that it does not use the alignment but the full reads [@garrison2012, @howtof].
 
-In this work, `bcftools` and `freebayes` variant callers were used to obtain biallelic SNPs and results were compared in terms of the number of SNPs, the overlap between sets, the depth at each position, and their minor allele frequencies.
-
-Best method could depend on nature of the sequencing data: eg. whole-genome, RAD-seq, RNA-seq, scRNA-seq, etc, and the depth of coverage.
+In this work, I used `bcftools` and `freebayes`callers to find biallelic SNPs.
+The results were compared in terms of the number of SNPs, the overlap between sets, the depth at each position, and their minor allele frequencies.
+Previous comparisons of variant callers have often found low concordance between tools and issues with sensitivity [@orawe2013, @cornish2015], though others have reported high concordance and accuracy in their results [@hwang2015].
+Best method could depend on nature of the sequencing data: eg.
+whole-genome, RAD-seq, RNA-seq, scRNA-seq, etc, and the depth of coverage.
 
@@
 
+-   Mean depth for each individual across all loci
+
+```{r vcf_indv_depth, engine = 'bash', eval = FALSE, echo = TRUE}
+vcftools --gzvcf mpileup.filter.recode.vcf.gz --depth --out mpileup
+vcftools --gzvcf freebayes.filter.recode.vcf.gz --depth --out freebayes
+```
+
+<!-- ## Relatedness -->
+
+<!-- vcftools --gzvcf mpileup.filter.recode.vcf.gz --relatedness --out mpileup -->
+
+<!-- vcftools --gzvcf freebayes.filter.recode.vcf.gz --relatedness --out freebayes -->
+
+<!-- ## Convert to (0,1,2) format -->
+
+<!-- vcftools --gzvcf mpileup.filter.recode.vcf.gz --012 --out mpileup -->
+
+<!-- vcftools --gzvcf freebayes.filter.recode.vcf.gz --012 --out freebayes -->
+
+<!-- quality, as shown in class, using what you think is a reasonable threshold.
+Compare the resulting vcf files in several ways.
+First, *compare the number of SNPs* and the *overlap of these SNPs across files*.
+-->
+
+<!-- Then, plot the *minor allele frequency* and depth of all SNPs (summed across individuals), and discuss the variation across loci. Compare across vcfs produced by different variant callers. -->
+
 ```{=html}
@@
 
-```{r load_data, message=FALSE}
+```{r load data, message=FALSE}
 #! {R code}
@@
   distinct() %>% 
-  bind_rows(tibble(SNPs = c('mpileup', 'freebayes'), 
+  bind_rows(
+    tibble(SNPs = c('mpileup', 'freebayes'), 
            Count = c(.[1,2][[1]] + .[3,2][[1]], 
@@
   arrange(desc(Count))
-```
 
-```{r load_data2, message=FALSE, echo=FALSE}
 # Allele frequencies
@@
   bind_rows(
+    read_tsv('./data/freebayes.filter.frq', col_names = cnames, skip = 1L) %>% 
+      mutate(caller = 'freebayes'),
     read_tsv('./data/mpileup.filter.frq', col_names = cnames, skip = 1L) %>% 
-      mutate(caller = 'mpileup'),
-    read_tsv('./data/freebayes.filter.frq', col_names = cnames, skip = 1L) %>% 
-      mutate(caller = 'freebayes')
+      mutate(caller = 'mpileup')
   ) %>% 
@@
   left_join(AlleleFreq, by = c('chrom', 'pos', 'caller'))
+# glimpse(vcf_df)
 
 cnames <- c('id', 'quality', 'n_snps', 'ts', 'tv', 'indels')
-qual_df <- bind_rows(
+qual_df <- 
+  read_tsv("./data/freebayes.quality.stats",
+                 skip = 1L, col_names = cnames) %>% 
+  mutate(caller = "freebayes") %>% 
+  bind_rows(
     read_tsv("./data/mpileup.quality.stats",
                  skip = 1L, col_names = cnames) %>% 
-  mutate(caller = "mpileup"),
-  read_tsv("./data/freebayes.quality.stats",
-                 skip = 1L, col_names = cnames) %>% 
-  mutate(caller = "freebayes")
-  ) %>% 
-  mutate(caller = as_factor(caller))
+  mutate(caller = "mpileup")
+  )
 
@@
 ```{r figs, echo=F}
-## this chunk contains code for all the plots
 library(ggbeeswarm)
@@
 library(grid)
-
 # Allele frequency histogram (alleles stacked)
@@
   ggplot(SiteDepth, aes(x = allele_depth, fill = caller, color = caller)) +
-  geom_density(alpha = 0.2) +
+  geom_density(alpha = 0.12) +
   geom_vline(aes(xintercept = mean(allele_depth), color = caller),
@@
   scale_x_log10() +
-  labs(x = 'depth per locus\n(sum of 10 individuals)', y = 'density') +
-  scale_color_carto_d(palette = 2, direction = -1) +
-  scale_fill_carto_d(palette = 2, direction = -1) +
+  labs(x = 'depth per locus', y = 'density') +
   guides(colour = guide_legend(override.aes = list(size=1, pch=15))) +
@@
   theme_bw()
-
-# Quality score distribution by caller fig
-fig_quality_dist <- qual_df %>% 
-  mutate(QUAL = round(quality/25)*25) %>% 
-  group_by(QUAL, caller) %>% 
-  summarise(n_snps = sum(n_snps),
-            ts  = sum(ts),
-            tv = sum(tv)) %>% 
-  pivot_longer(cols = c(ts,tv)) %>% 
-  ggplot(aes(x = QUAL, y = value, fill = name)) +
-  geom_col(width = 19) +
-  facet_wrap(~caller) +
-  scale_fill_carto_d(palette = 2) +
-  labs(x="quality score", y = "loci", fill = NULL) +
-  scale_y_log10() +
-  theme_bw()
-
 ```
@@
 
-```{r fig1, echo=F, fig.align = 'center', fig.width=8, fig.height=3.5, fig.cap="(A) Distribution of minor allele frequency; SNPs with frequency = 0 or 1 were omitted.
+```{r fig1, echo=F, fig.align = 'center', fig.width=7, fig.height=3, fig.cap="(A) Distribution of minor allele frequency; SNPs with frequency = 0 or 1 were omitted.
 (B) Distribution of depth per locus summed across all individuals (note log-scaled x-axis)."}
@@
 plot_annotation(tag_levels = 'A')
+```
+
+```{=html}
+<!-- Zeggini, E., & Morris, A. (2015). Assessing Rare Variation in Complex Traits Design and Analysis of Genetic Studies (1st ed. 2015.). Springer 
+New York. <https://doi.org/10.1007/978-1-4939-2824-8> -->
+```
+```{r engine = 'bash', eval=FALSE}
+## quality scores...
+grep -E "Quality" mpileup.filter.recode.stats | cut -f1 --complement > mpileup.quality.stats
+grep ^QUAL mpileup.filter.recode.stats | cut -f1 --complement >> mpileup.quality.stats 
+grep -E "Quality" freebayes.filter.recode.stats | cut -f1 --complement > freebayes.quality.stats
+grep ^QUAL freebayes.filter.recode.stats | cut -f1 --complement >> freebayes.quality.stats 
+```
+
+```{r echo = F}
+fig_quality_dist <- qual_df %>% 
+  mutate(QUAL = round(quality/25)*25) %>% 
+  group_by(QUAL, caller) %>% 
+  summarise(n_snps = sum(n_snps),
+            ts  = sum(ts),
+            tv = sum(tv)) %>% 
+  pivot_longer(cols = c(ts,tv)) %>% 
+  ggplot(aes(x = QUAL, y = value, fill = name)) +
+  geom_col(width = 19) +
+  facet_wrap(~caller) +
+  scale_fill_carto_d(palette = 2) +
+  labs(x="quality score", y = "loci", fill = NULL) +
+  scale_y_log10() +
+  theme_bw()
+
 ```
@@
 
-```{r fig_allele_maf_nchr, echo=F, fig.align='center', fig.height=4, fig.cap="Allele depth and minor allele frequency do not show any correlation; minor allele frequency is constrained to specific values by the number of chromosomes represented"}
+```{r}
 fig_alleledepth_over_maf
@@
 
-## Mean depth by subject
-
-Mean depth for each individual across all loci was computed as follows:
-
-```{r vcf_indv_depth, engine = 'bash', eval = FALSE, echo = TRUE}
-vcftools --gzvcf mpileup.filter.recode.vcf.gz --depth --out mpileup
-vcftools --gzvcf freebayes.filter.recode.vcf.gz --depth --out freebayes
-```
-
-We can see the all individuals have greater mean depth at SNPs called by `freebayes` than `bcftools mpileup` (fig. 4)
-
-```{r fig_indv_depth, echo=F, fig.align = 'center', fig.width=5, fig.height=2.9, fig.cap="Mean depth per locus for each individual is greater for SNPs called by  freebayes than bcftools."}
+```{r fig2, echo=F, fig.align = 'center', fig.width=5, fig.height=2.9, fig.cap="Mean depth per locus for each individual."}
 fig_indvdepth
@@
 
-### Quality scores
-
-I obtained quality scores for each loci from bctools stats
-
-```{r engine = 'bash', eval=FALSE}
-## quality scores...
-bcftools stats mpileup.filter.recode.vcf > mpileup.filter.recode.stats
-grep -E "Quality" mpileup.filter.recode.stats | cut -f1 --complement > mpileup.quality.stats
-grep ^QUAL mpileup.filter.recode.stats | cut -f1 --complement >> mpileup.quality.stats bcftools stats freebayes.filter.recode.vcf > freebayes.filter.recode.stats
-grep -E "Quality" freebayes.filter.recode.stats | cut -f1 --complement > freebayes.quality.stats
-grep ^QUAL freebayes.filter.recode.stats | cut -f1 --complement >> freebayes.quality.stats
-```
-
-```{r fig3, echo=F, fig.align = 'center', fig.width=5, fig.height=2.9, fig.cap="Minor allele fre"}
+```{r fig3, echo=F, fig.align = 'center', fig.width=5, fig.height=2.9, fig.cap="Mean depth per locus for each individual."}
 fig_alleledepth_over_maf
@@
 
-```{r fig4, echo=F, fig.height = 3, fig.align = 'center', fig.width=5, fig.height=2.9, fig.cap="The quality score distributions of freebayes and bcftools mpileup "}
-fig_quality_dist
+```{r}
+
 ```
-
-<!-- ## Relatedness -->
-
-<!-- vcftools --gzvcf mpileup.filter.recode.vcf.gz --relatedness --out mpileup -->
-
-<!-- vcftools --gzvcf freebayes.filter.recode.vcf.gz --relatedness --out freebayes -->
-
-<!-- ## Convert to (0,1,2) format -->
-
-<!-- vcftools --gzvcf mpileup.filter.recode.vcf.gz --012 --out mpileup -->
-
-<!-- vcftools --gzvcf freebayes.filter.recode.vcf.gz --012 --out freebayes -->
-
-<!-- quality, as shown in class, using what you think is a reasonable threshold.
-Compare the resulting vcf files in several ways.
-First, *compare the number of SNPs* and the *overlap of these SNPs across files*.
--->
-
-<!-- Then, plot the *minor allele frequency* and depth of all SNPs (summed across individuals), and discuss the variation across loci. Compare across vcfs produced by different variant callers. -->
 
@@
 # References
-
-```{=html}
-<!-- Zeggini, E., & Morris, A. (2015). Assessing Rare Variation in Complex Traits Design and Analysis of Genetic Studies (1st ed. 2015.). Springer 
-New York. <https://doi.org/10.1007/978-1-4939-2824-8> -->
-```

### 2021-03-18 manuscript.Rmd
not even sure what has changed

@@
 In the current work I compared variant calling by two tools: `bcftools mpileup/call` (previously `samtools`) and `freebayes`.
+The results were compared in terms of the number of SNPs, the overlap between sets, the depth at each position, and their minor allele frequencies.
 Both `bcftools` and `freebayes` apply Bayesian models to estimate genotype posterior probabilities, though they differ in how this is implemented.
@@
 `freebayes` a uses different approach to inferring haplotypes in that it does not use the alignment but the full reads [@garrison2012, @howtof].
-
-In this work, I used `bcftools` and `freebayes`callers to find biallelic SNPs.
-The results were compared in terms of the number of SNPs, the overlap between sets, the depth at each position, and their minor allele frequencies.
-Previous comparisons of variant callers have often found low concordance between tools and issues with sensitivity [@orawe2013, @cornish2015], though others have reported high concordance and accuracy in their results [@hwang2015].
-Best method could depend on nature of the sequencing data: eg.
-whole-genome, RAD-seq, RNA-seq, scRNA-seq, etc, and the depth of coverage.
+Previous comparisons of variant callers have often found low concordance between tools and issues with sensitivity [@orawe2013, @cornish2015].
+However, others have reported high concordance and accuracy in their evaluation [@hwang2015]; the best method is likely dependent on nature of the sequencing data: (whole-genome or exome, RAD-seq, RNA-seq, *etc*.), the depth of coverage, and the type of variants of interest (*eg.* SNPs or indels).
 

### 2021-03-18 manuscript.Rmd
header change

@@
 ---
-title: 'Project 3: Variant calling in Burbot (*Lota lota*) using `freebayes` and `mpileup`'
-author: "J Moggridge"
-date: "06/03/2021"
+title: ''
+author: ""
 output: 
@@
 ---
+
+# Project 3: Variant calling in Burbot (*Lota lota*) using `freebayes` and `mpileup`
+
+Jason Moggridge 1159229
 
@@
 
-<!-- purpose of variant calling -->
-
-The analysis of genetic variants (single-nucleotide polymorphisms -SNPs and insertions/deletions - indels), by high-throughput sequencing has applications in studies of phylogeny and population structure, quantitative-trait loci, genome-wide associations, among others.
-Such polymorphic sites are also of practical interest in medical genomics, *eg*.
+The analysis of genetic variants (single-nucleotide polymorphisms -SNPs and insertions/deletions - indels) by high-throughput sequencing has applications in studies of phylogeny and population structure, quantitative-trait loci, genome-wide associations, among others.
+Such polymorphic sites are also of practical interest in clinical genomics, *eg*.
 to discover disease-causing mutations [@cornish2015], and in agriculture, where SNPs can be used as genetic markers to guide breeding of crops [@yao2020] and livestock [@ni2015].
@@
 
-------------------------------------------------------------------------
-
 # Methods
-
-Data can be found on `cedar` in `/scratch/jmoggrid/Project3/`. <!-- (chmod 755 /scratch/username/directory/ -R). --> <!-- srun --pty --account="def-nricker" -t 0-03:00:00 --mem=32000 /bin/bash -->
 
@@
 
+*All files can be found on `cedar` in `/scratch/jmoggrid/Project3/`.*
+
 I used data from project \#2 in this work, specifically the reads mapped to the burbot reference with the `bowtie2` aligner.
 I used `bcftools mpileup` and `GATK HaplotypeCaller` to call variants for the ten individuals using these alignments.
+I first needed to convert my `.sam` files to `.bam` format and index them, as well as adding read groups using `picard` (necessary for `freebayes`).
 
-I first needed to convert my `.sam` files to `.bam` format and index them, as well as adding read groups using `picard` (necessary for `freebayes`).
+<!-- fyi - bash code chunks have engine='bash' in their header -->
 

### 2021-03-19 manuscript.Rmd
wrote a discussion, of sorts

@@
 
-# Project 3: Variant calling in Burbot (*Lota lota*) using `freebayes` and `mpileup`
+# Project 3: Comparison of variant calling by `freebayes` and `mpileup` from high-throughput sequencing of Burbot (*Lota lota*)
 
@@
 This implementation has the advantage of being very fast.
-`freebayes` a uses different approach to inferring haplotypes in that it does not use the alignment but the full reads [@garrison2012, @howtof].
+`freebayes` a uses different approach to inferring haplotypes in that it does not use the alignment but the full reads [@garrison2012; @howtof].
 Previous comparisons of variant callers have often found low concordance between tools and issues with sensitivity [@orawe2013, @cornish2015].
@@
 
+*All files can be found on `cedar` in `/scratch/jmoggrid/Project3/`.*
+
 ### Dataset
 
-*All files can be found on `cedar` in `/scratch/jmoggrid/Project3/`.*
-
 I used data from project \#2 in this work, specifically the reads mapped to the burbot reference with the `bowtie2` aligner.
-I used `bcftools mpileup` and `GATK HaplotypeCaller` to call variants for the ten individuals using these alignments.
-I first needed to convert my `.sam` files to `.bam` format and index them, as well as adding read groups using `picard` (necessary for `freebayes`).
+First, I needed to convert my `.sam` files to `.bam` format and index them, as well as adding read groups using `picard` (necessary for `freebayes`):
 
@@
 
-I ran `bcftools` variant calling for the 10 .sorted.bam files.
-For `mpileup`, I added arguments to provide annotations for read depth and allelic depth (-a DP,AD) and to specify Illumina as the sequencing platform (-P).
-For `call`, I provided arguments to specify the use of the multi-allelic caller, 'variants only', to skip indels, and to provide the genotype quality scores (GQ).
+I used `bcftools mpileup/call` (referred to as `mpileup`) and `freebayes` to call variants from the ten burbot alignments.
+For `mpileup`, I added arguments to include annotations for read depth and allelic depth (-a DP,AD) and to specify Illumina as the sequencing platform (-P).
+For `call`, I specified the use of the newer multi-allelic caller (m), 'variants only' (v), to skip indels, and to provide the genotype quality scores (GQ).
+A summary of the raw SNP calls was obtained using `bcftools stats`.
 
@@
 
-I called SNPs using `freebayes` with default settings; annotations for genotype quality were requested with `--genotype-qualities`. Output was limited to SNPs by piping to `vcffilter`.
+I called SNPs using `freebayes` with default settings except for the flag to provide annotations for genotype quality. Output was limited to SNPs with `vcffilter`.
 
@@
 I used the `vcftools` library to filter variant calls.
-Biallelic SNPs (`--min-alleles 2 --max-alleles 2 --remove-indels`) were retained if they had a quality score of at least 20 (`--minQ 30`).
+Biallelic SNPs (`--min-alleles 2 --max-alleles 2 --remove-indels`) were retained if they had a quality score of at least 20 (`--minQ 20`).
 Genotypes were retained at these sites if they had genotyping quality score of at least 10 (`--minGQ 10`).
-I found the vast majority of variants to be shared among all 10 individuals, so I applied filters for the maximum and minimum minor allele frequencies, such that these would not be either zero or one (`–maf 0.01 –max-maf 0.99`).
+I found the vast majority of variants to be shared among all 10 individuals (data not shown), so I applied filters for the maximum and minimum minor allele frequencies, such that these would not be either zero or one (`–maf 0.01 –max-maf 0.99`).
 Any calls that did not receive a 'PASS' flag were discarded (`--remove-filtered-all`).
@@
 bcftools isec -p isec mpileup.filter.recode.vcf.gz freebayes.filter.recode.vcf.gz
+# Get overlap data from isec into tsv format
+> Overlap.vals
+for file in isec/*.vcf
+do
+  bcftools stats $file | grep "number of SNPs:" | cut -f4 >> Overlap.vals
+done
+echo $'unique to mpileup\nunique to freebayes\nshared\nshared' > Overlap.names
+paste  Overlap.names Overlap.vals > Overlap.tsv
+sed -i '1s/^/SNPs\tCount\n/' Overlap.tsv
 ```
 
-<!-- # Get overlap data from isec into tsv format -->
-
-<!-- > Overlap.vals -->
-
-<!-- for file in isec/*.vcf -->
-
-<!-- do  -->
-
-<!--   bcftools stats $file| grep "number of SNPs:" | cut -f4 >> Overlap.vals -->
-
-<!-- done -->
-
-<!-- echo $'unique to mpileup\nunique to freebayes\nshared\nshared' > Overlap.names -->
-
-<!-- paste  Overlap.names Overlap.vals > Overlap.tsv -->
-
-<!-- sed -i '1s/^/SNPs\tCount\n/' Overlap.tsv -->
-
-Using `vcftools`, I computed the following:
-
--   Allele frequency at each loci across individuals
+Using `vcftools`, I computed the allele frequency at each loci across individuals, as well as the depth at each locus summed across individuals.
 
 ```{r vcf_freq, engine = 'bash', eval = FALSE, echo = TRUE}
-vcftools --gzvcf mpileup.filter.recode.vcf.gz --freq \
-  --out mpileup.filter
+# allele frequency
+vcftools --gzvcf mpileup.filter.recode.vcf.gz --freq  --out mpileup.filter
 vcftools --gzvcf freebayes.filter.recode.vcf.gz --freq --out freebayes.filter
-```
-
--   Sum depth at each locus across individuals
-
-```{r vcf_site_depth, engine = 'bash', eval = FALSE, echo = TRUE}
+# site depth sum
 vcftools --gzvcf mpileup.filter.recode.vcf.gz --site-depth --out mpileup
@@
 
--   Mean depth for each individual across all loci
-
-```{r vcf_indv_depth, engine = 'bash', eval = FALSE, echo = TRUE}
-vcftools --gzvcf mpileup.filter.recode.vcf.gz --depth --out mpileup
-vcftools --gzvcf freebayes.filter.recode.vcf.gz --depth --out freebayes
-```
-
-<!-- ## Relatedness -->
-
-<!-- vcftools --gzvcf mpileup.filter.recode.vcf.gz --relatedness --out mpileup -->
-
-<!-- vcftools --gzvcf freebayes.filter.recode.vcf.gz --relatedness --out freebayes -->
-
-<!-- ## Convert to (0,1,2) format -->
-
-<!-- vcftools --gzvcf mpileup.filter.recode.vcf.gz --012 --out mpileup -->
-
-<!-- vcftools --gzvcf freebayes.filter.recode.vcf.gz --012 --out freebayes -->
-
-<!-- quality, as shown in class, using what you think is a reasonable threshold.
-Compare the resulting vcf files in several ways.
-First, *compare the number of SNPs* and the *overlap of these SNPs across files*.
--->
-
-<!-- Then, plot the *minor allele frequency* and depth of all SNPs (summed across individuals), and discuss the variation across loci. Compare across vcfs produced by different variant callers. -->
-
-```{=html}
-<!-- # get data down from cedar in local terminal:
- scp jmoggrid@cedar.computecanada.ca:/scratch/jmoggrid/Project3/Overlap.tsv .
- scp jmoggrid@cedar.computecanada.ca:/scratch/jmoggrid/Project3/*.frq . 
- scp jmoggrid@cedar.computecanada.ca:/scratch/jmoggrid/Project3/*.idepth . 
- scp jmoggrid@cedar.computecanada.ca:/scratch/jmoggrid/Project3/*.ldepth .  -->
-```
-
-------------------------------------------------------------------------
-
 ### Statistics
 
-I created figures and summary tables from the data generated above using the `R` language.
+I created figures and summary tables from the data generated above using the `R` language. *This code is presented in the supplementary section at the end of the manuscript.*
 
@@
 
-# Overlap counts table; add
+# Overlap counts table; add totals for freebayes and mpileup 
+Overlap <- read_delim('./data/Overlap.tsv', '\t') %>%  
+  distinct() %>% 
+  bind_rows(
+    tibble(SNPs = c('mpileup', 'freebayes'), 
+           Count = c(.[1,2][[1]] + .[3,2][[1]], 
+                     .[2,2][[1]] + .[3,2][[1]]))) %>%  
+  arrange(desc(Count))
+```
+
+```{r load_data2, message=FALSE, echo = F}
+# Allele frequencies
+cnames <- c('chrom','pos', 'n_alleles','n_chr','a1','a2')
+AlleleFreq <-  
+  bind_rows(
+    read_tsv('./data/freebayes.filter.frq', col_names = cnames, skip = 1L) %>% 
+      mutate(caller = 'freebayes'),
+    read_tsv('./data/mpileup.filter.frq', col_names = cnames, skip = 1L) %>% 
+      mutate(caller = 'mpileup')
+  ) %>% 
+  mutate(allele = str_extract(a1, '[A-Z]+'), 
+         frequency = str_remove_all(a1, '[A-Z]+:'),
+         allele2 = str_extract(a2, '[A-Z]+'), 
+         frequency2 = str_remove_all(a2, '[A-Z]+:')
+  ) %>% 
+  mutate(across(contains('freq'), as.numeric)) %>%
+  mutate(
+    maf = ifelse(frequency > 0.5, 1 - frequency, frequency),
+    minor_allele = ifelse(maf == frequency, allele, allele2)
+  )
+
+# Depth
+cnames <- c('individual', 'n sites', 'mean_depth')
+IndvDepth <- bind_rows(
+    read_tsv('./data/mpileup.idepth', col_names = cnames, skip = 1L) %>%
+      mutate(caller = 'mpileup'),
+    read_tsv('./data/freebayes.idepth', col_names = cnames, skip = 1L) %>%
+      mutate(caller = 'freebayes')
+  ) %>% 
+  mutate(individual = str_remove_all(individual, 'bam.|.bowtie.burbot'),
+         caller = as_factor(caller)) %>% 
+  arrange(individual)
+
+# Site Depth
+cnames <- c("chr", "pos", "allele_depth", "var_depth")
+SiteDepth <- bind_rows(
+  read_tsv('./data/mpileup.ldepth', col_names = cnames, skip = 1L) %>%
+    mutate(caller = 'mpileup'),
+  read_tsv('./data/freebayes.ldepth', col_names = cnames, skip = 1L) %>%
+    mutate(caller = 'freebayes')
+) %>% mutate(caller = as_factor(caller))
+
+vcf_df <- SiteDepth %>%
+  rename(chrom = chr) %>% 
+  left_join(AlleleFreq, by = c('chrom', 'pos', 'caller'))
+
+cnames <- c('id', 'quality', 'n_snps', 'ts', 'tv', 'indels')
+qual_df <- 
+  bind_rows(
+    read_tsv("./data/mpileup.quality.stats", skip = 1L, col_names = cnames) %>%  mutate(caller = "mpileup"),
+    read_tsv("./data/freebayes.quality.stats", skip = 1L, col_names = cnames) %>%  mutate(caller = "freebayes")
+  ) %>%  mutate(caller = as_factor(caller))
+```
+
+```{r figs, echo=F}
+library(ggbeeswarm)
+library(patchwork)
+library(rcartocolor)
+library(gridExtra)
+library(grid)
+
+# Allele frequency histogram (alleles stacked)
+fig_allelefreq <- 
+  ggplot(AlleleFreq) +
+  geom_histogram(aes(x = maf), bins = 20, alpha = 0.8) +
+  facet_wrap(caller~., scales = 'free_y') +
+  labs(x = 'minor allele frequency', y='loci') +
+  theme_bw()
+
+# The distribution of depth per-site summed across individuals
+fig_sitedepth <- 
+  ggplot(SiteDepth, aes(x = allele_depth, fill = caller, color = caller)) +
+  geom_density(alpha = 0.12) +
+  geom_vline(aes(xintercept = mean(allele_depth), color = caller),
+             lty = 2) +
+  scale_x_log10() +
+  labs(x = 'depth per locus', y = 'density') +
+  guides(colour = guide_legend(override.aes = list(size=1, pch=15))) +
+  theme_bw() +
+  theme(legend.position = c(.8,.7))
+
+## SUPPLEMENTAL
+
+# Allele depth over maf, coloured by n chromosomes
+fig_alleledepth_over_maf <- 
+  ggplot(vcf_df, aes(x=maf, y=allele_depth, color = n_chr)) +
+  geom_point(size = 0.5, alpha = 0.7) +
+  scale_y_log10() +
+  facet_wrap(~caller) +
+  scale_color_viridis_c() +
+  labs(x = "Minor allele frequency",
+       y = "Allele depth", 
+       color="Number of\nchromosomes") +
+  theme_bw()
+
+# The distribution of mean depth per site within individuals
+set.seed(45)
+fig_indvdepth <-
+  ggplot(IndvDepth, aes(y = mean_depth, x = caller)) +
+  geom_violin(alpha = 0.1, draw_quantiles = c(.25, .5, .75), trim = F) +
+  geom_quasirandom(aes(color = individual), size = 3, width = 0.3) +
+  scale_color_carto_d(palette = 2) +
+  guides(colour = guide_legend(override.aes = list(size = 4, pch = 16))) +
+  facet_wrap( ~ caller, scales = 'free_x') +
+  labs(y = 'mean depth per locus', x = '', color = 'Individual') +
+  theme_bw() +
+  theme(
+    axis.text.x = element_blank(),
+    axis.ticks.x = element_blank(),
+    legend.position = 'right'
+  )
+fig_quality_dist <- qual_df %>% 
+  mutate(QUAL = round(quality/25)*25) %>% 
+  group_by(QUAL, caller) %>% 
+  summarise(n_snps = sum(n_snps),
+            ts  = sum(ts),
+            tv = sum(tv)) %>% 
+  pivot_longer(cols = c(ts,tv)) %>% 
+  ggplot(aes(x = QUAL, y = value, fill = name)) +
+  geom_col(width = 19) +
+  facet_wrap(~caller) +
+  scale_fill_carto_d(palette = 2) +
+  labs(x="quality score", y = "loci", fill = NULL) +
+  scale_y_log10() +
+  theme_bw()
+```
+
+------------------------------------------------------------------------
+
+\newpage
+
+# Results
+
+From the ten burbots in our sample, `mpileup` called a total of 132,046 SNPs, 173 of which were multiallelic, and with a transition:transversion ratio (ts:tv) of 0.81; `freebayes`called a total of 106,973 SNPs, with 1761 being multiallelic, and ts:tv of 0.44 (table 1).
+After applying filters for these were reduced to 11,115 SNPs with a 0.87 transition:transversion ratio for `mpileup` calls and 6,722 SNPs with 1.09 transition:transversion ratio for `freebayes` calls.
+The sets of biallelic SNPs identified by each caller were compared (table 2): `mpileup` called `r round(Overlap[1,2][[1]] * 100 / Overlap[2,2][[1]], 2) - 100` % more biallelic SNPs than `freebayes`.
+Of these, a large proportion (`r scales::comma(Overlap[3,2][[1]])`) were shared between both sets, representing `r round(Overlap[3,2][[1]] * 100 /Overlap[1,2][[1]], 2)` % of `mpileup` calls and `r round(Overlap[3,2][[1]] * 100 /Overlap[2,2][[1]], 2)` % of `freebayes` calls.
+Additionally, `r Overlap[4,2][[1]]` SNPs were unique to `mpileup` and `r Overlap[5,2][[1]]` to `freebayes`.
+
+The distribution minor allele frequencies of SNPs from `mpileup` shows a U-shape with peaks at 0.05 and 0.5, whereas those from `freebayes` peak at 0.05 and are otherwise evenly distributed (fig.1A).
+The range of possible values is constrained due to the small sample size (some values can only be obtained with a specific amount of missing data; fig.
+2A).
+The sum depth per locus across all individuals is generally in the range of 50-200 reads, with both callers showing similar distributions (fig.
+1B).
+The mean depth for `mpileup` was 173.7 reads and for `freebayes` 162.2 reads.
+Both callers have some extreme outliers in terms of site depth: the maximum depth for `freebayes` was 16,162 reads, while for `mpileup` it was only 2522 reads.
+The minimum site depth was 5 and 3 for `freebayes` and `mpileup` respectively and `mpileup` has a greater proportion of low-depth sites in general.
+Freebayes also had a greater mean depth per individual (fig.
+2B).
+Both callers show similar distributions of quality scores, though `mpileup` calls a greater proportion of transversions at low-quality sites (fig.
+2C).
+
+------------------------------------------------------------------------
+
+```{r table1_filtering, echo=FALSE}
+tribble(
+  ~ Caller, ~ `raw SNPs`, ~ `raw ts/tv`, ~ `post-filter SNPs`, ~ `post-filter ts/tv`,
+  "mpileup", 132046, 0.81, 11115, 0.87,
+  "freebayes", 106973, 0.44, 6722, 1.09
+) %>% pander::pander(caption = "Raw and filtered SNP counts and transition/transversion ratios for calls by mpileup and freebayes.")
+```
+
+```{r fig1, echo=F, fig.align = 'center', fig.width=7, fig.height=3, fig.cap="(A) Distribution of minor allele frequency; SNPs with frequency = 0 or 1 were omitted.
+(B) Distribution of depth per locus summed across all individuals (note log-scaled x-axis)."}
+(fig_allelefreq + fig_sitedepth) &
+  plot_layout(guides = 'keep') &
+  plot_annotation(tag_levels = 'A')
+```
+
+```{r table2_overlap, echo=F}
+Overlap %>% 
+  pander::pander(caption = "Summary of total, shared, and unique SNPs called by mpileup and freebayes (after filtering).")
+```
+
+```{r fig2, echo=F, fig.align = 'center', fig.width=6, fig.height=7, fig.cap="(A) Scatterplot of loci (points) by allele depth and minor allele frequency, colored by the number of chromosomes counted.
+Due to the sample size (n=10), minor allele frequencies are constrained by the number of chromosomes at the SNP site; there appears to be no relationship between allele depth (ie.
+read coverage) and minor allele frequencies.
+\n(B) Mean depth per locus for each individual.
+Violin plot lines indicate quartiles.
+\n(C) The quality score distributions for each set of SNP calls, separated into transitons (ts) and transversions (tv)"}
+fig_alleledepth_over_maf / fig_indvdepth / fig_quality_dist & 
+  plot_annotation(tag_levels = 'A')
+```
+
+# Discussion
+
+In this work I compared the sets of biallelic SNPs called by two tools, bcftools mpileup/call and freebayes, from alignments of burbot reads from 10 individuals.
+The raw SNP calls were filtered similarly for both sets based on quality scores, genotyping confidence scores, and minor allele frequency; most of the data (\>90%) was discarded during filtering (table 1), with roughly one-third of calls having allele frequency of 0 or 1 (*ie.* all individuals were similar but differed from the reference).
+Unexpectedly, I found that mpileup called many more SNPs that freebayes, and retained nearly twice as many after filtering.
+This is in contrast to findings from @cornish2015 who found that freebayes called more SNPs in human exome data, though both tools had similar-sized sets after filtering.
+The discrepancy between these findings does not seem to be explained by the slight differences in filtering (they filtered for read depth; I filtered for genotype quality), but could be related to differences between datasets [@hwang2015].
+
+Interestingly, the sets of SNPs called by mpileup and freebayes showed a high degree of overlap, with over 90% of freebayes calls being shared with the larger mpileup set (table 2).
+This result is somewhat in agreement with @hwang2015, who found \~90% concordance between (Samtools) mpileup, freebayes, and GATK on the gold-standard Genome in a Bottle dataset.
+My results suggest then, that either mpileup has low specificity (calls variants that don't exist), or that freebayes has low sensitivity (misses true variants).
+This does not seem to be the case generally though, as @cornish2015 suggest the opposite and found that a variety of tools had similarly low sensitivity on the GiaB gold-standard data.
+
+My analysis also revealed differences in the distribution of minor allele frequencies (MAF) and read depth at SNP loci (fig.
+1).
+Specifically, mpileup shows a clear preference for SNPs with MAF of 0.5 and to a lesser degree 0.25 (relative to similar values).
+Both callers have a peak at low MAF (0.05) but otherwise mpileup has an even distribution from 0.10 to 0.5, when accounting for the constraint on these values by the small sample size (fig.
+2A).
+The callers show differences in the read depth across loci (fig.
+1B) and individuals (fig.
+2B), as high-depth sites (\>250/ind.) are hard-filtered by `mpileup`, and mpileup identifies many more low-depth sites (\<70x) that are ignored by freebayes.
+As such, the concordance between sets would likely increase if stricter filters were used to mitigate these biases.
+
+Certain characteristics of the burbot data present challenges for variant calling.
+The highly-fragmented assembly means we are likely to miss many variants since they are not represented in the alignment.
+The relatively low coverage of the data means that a number of true variants are likely discarded because of high uncertainty in their classification.
+Further, burbot is not a model organism and does not have the same resources to develop useful prior probabilities for variant calling (eg.
+SNP databases for human variant calling).
+In the present work, the small number of subjects means that the identification of uncommon and rare variants is highly unlikely.
+If we were more interested in finding such loci, we would want to apply less stringent filtering and accept the cost of increased false discoveries.
+
+------------------------------------------------------------------------
+
+\newpage
+
+# Supplemental
+
+I also examined the quality score distribution over sites and the mean per site depth by individual. These were computed with `bcftools stats` and `vcftools` respectively, as follows:
+
+```{r engine = 'bash', eval=FALSE}
+## quality scores from bf
+grep -E "Quality" mpileup.filter.recode.stats | cut -f1 --complement > mpileup.quality.stats
+grep ^QUAL mpileup.filter.recode.stats | cut -f1 --complement >> mpileup.quality.stats 
+grep -E "Quality" freebayes.filter.recode.stats | cut -f1 --complement > freebayes.quality.stats
+grep ^QUAL freebayes.filter.recode.stats | cut -f1 --complement >> freebayes.quality.stats 
+# Mean depth for each individual across all loci
+vcftools --gzvcf mpileup.filter.recode.vcf.gz --depth --out mpileup
+vcftools --gzvcf freebayes.filter.recode.vcf.gz --depth --out freebayes
+```
+
+### R code for parsing results data
+
+```{r supplementary1, eval = F, echo=T, include=T}
+#! {R code}
+library(tidyverse, quietly = T)
+library(ggbeeswarm)
+library(patchwork)
+library(rcartocolor)
+library(gridExtra)
+library(grid)
+
+# Overlap counts table; add totals for freebayes and mpileup 
 Overlap <- read_delim('./data/Overlap.tsv', '\t') %>%  
@@
   left_join(AlleleFreq, by = c('chrom', 'pos', 'caller'))
-# glimpse(vcf_df)
 
@@
 qual_df <- 
-  read_tsv("./data/freebayes.quality.stats",
-                 skip = 1L, col_names = cnames) %>% 
-  mutate(caller = "freebayes") %>% 
   bind_rows(
-    read_tsv("./data/mpileup.quality.stats",
-                 skip = 1L, col_names = cnames) %>% 
-  mutate(caller = "mpileup")
-  )
-
+    read_tsv("./data/mpileup.quality.stats", skip = 1L, col_names = cnames) %>%  mutate(caller = "mpileup"),
+    read_tsv("./data/freebayes.quality.stats", skip = 1L, col_names = cnames) %>%  mutate(caller = "freebayes")
+  ) %>%  mutate(caller = as_factor(caller))
 ```
 
-```{r figs, echo=F}
-library(ggbeeswarm)
-library(patchwork)
-library(rcartocolor)
-library(gridExtra)
-library(grid)
+### R code for plotting results
+
+```{r supplementary2, eval = F, echo=T, include=T}
 # Allele frequency histogram (alleles stacked)
@@
 
+## SUPPLEMENTAL
+
+# Allele depth over maf, coloured by n chromosomes
+fig_alleledepth_over_maf <- 
+  ggplot(vcf_df, aes(x=maf, y=allele_depth, color = n_chr)) +
+  geom_point(size = 0.5, alpha = 0.7) +
+  scale_y_log10() +
+  facet_wrap(~caller) +
+  scale_color_viridis_c() +
+  labs(x = "Minor allele frequency",
+       y = "Allele depth", 
+       color="Number of\nchromosomes") +
+  theme_bw()
+
 # The distribution of mean depth per site within individuals
@@
   )
-
-# Allele depth over maf, coloured by n chromosomes
-fig_alleledepth_over_maf <- 
-  ggplot(vcf_df, aes(x=maf, y=allele_depth, color = n_chr)) +
-  geom_point(size = 0.5, alpha = 0.7) +
-  scale_y_log10() +
-  facet_wrap(~caller) +
-  scale_color_viridis_c() +
-  labs(x = "Minor allele frequency",
-       y = "Allele depth", 
-       color="Number of\nchromosomes") +
-  theme_bw()
-```
-
-------------------------------------------------------------------------
-
-\newpage
-
-# Results
-
-From the ten burbots in our sample, `mpileup` called a total of 132,046 SNPs, 173 of which were multiallelic, and with a transition:transversion ratio (ts:tv) of 0.81; `freebayes`called a total of 106,973 SNPs, with 1761 being multiallelic, and ts:tv of 0.44 (table 1).
-After applying filters for these were reduced to 11,115 SNPs with a 0.87 transition:transversion ratio for `mpileup` calls and 6,722 SNPs with 1.09 transition:transversion ratio for `freebayes` calls.
-The sets of biallelic SNPs identified by each caller were compared (table 2): `mpileup` called `r round(Overlap[1,2][[1]] * 100 / Overlap[2,2][[1]], 2) - 100` % more biallelic SNPs than `freebayes`.
-Of these, a large proportion (`r scales::comma(Overlap[3,2][[1]])`) were shared between both sets, representing `r round(Overlap[3,2][[1]] * 100 /Overlap[1,2][[1]], 2)` % of `mpileup` calls and `r round(Overlap[3,2][[1]] * 100 /Overlap[2,2][[1]], 2)` % of `freebayes` calls.
-Additionally, `r Overlap[4,2][[1]]` SNPs were unique to `mpileup` and `r Overlap[5,2][[1]]` to `freebayes`.
-
-The distribution minor allele frequencies of SNPs from `mpileup` shows a U-shape with peaks at 0.05 and 0.5, whereas those from `freebayes` peak at 0.05 and are otherwise evenly distributed (fig.1A).
-The sum depth per locus across all individuals is generally in the range of 50-200 reads, with both callers showing similar distributions (fig.1B).
-The mean depth for `mpileup` was 173.7 reads and for `freebayes` 162.2 reads.
-Both callers have some extreme outliers in terms of site depth: the maximum depth for `freebayes` was 16,162 reads, while for `mpileup` it was only 2522 reads.
-The minimum site depth was 5 and 3 for `freebayes` and `mpileup` respectively and `mpileup` has a greater proportion of low-depth sites.
-
-------------------------------------------------------------------------
-
-```{r table1_filtering, echo=FALSE}
-tribble(
-  ~ Caller, ~ `raw SNPs`, ~ `raw ts/tv`, ~ `post-filter SNPs`, ~ `post-filter ts/tv`,
-  "mpileup", 132046, 0.81, 11115, 0.87,
-  "freebayes", 106973, 0.44, 6722, 1.09
-) %>% pander::pander(caption = "Raw and filtered SNP counts and transition/transversion ratios for calls by mpileup and freebayes.")
-```
-
-```{r table2_overlap, echo=F}
-Overlap %>% pander::pander(caption = "Total, shared, and unique SNPs called by mpileup and freebayes (after filtering).")
-```
-
-```{r fig1, echo=F, fig.align = 'center', fig.width=7, fig.height=3, fig.cap="(A) Distribution of minor allele frequency; SNPs with frequency = 0 or 1 were omitted.
-(B) Distribution of depth per locus summed across all individuals (note log-scaled x-axis)."}
-
-(fig_allelefreq + fig_sitedepth) &
-plot_layout(guides = 'keep') &
-plot_annotation(tag_levels = 'A')
-```
-
-```{=html}
-<!-- Zeggini, E., & Morris, A. (2015). Assessing Rare Variation in Complex Traits Design and Analysis of Genetic Studies (1st ed. 2015.). Springer 
-New York. <https://doi.org/10.1007/978-1-4939-2824-8> -->
-```
-```{r engine = 'bash', eval=FALSE}
-## quality scores...
-grep -E "Quality" mpileup.filter.recode.stats | cut -f1 --complement > mpileup.quality.stats
-grep ^QUAL mpileup.filter.recode.stats | cut -f1 --complement >> mpileup.quality.stats 
-grep -E "Quality" freebayes.filter.recode.stats | cut -f1 --complement > freebayes.quality.stats
-grep ^QUAL freebayes.filter.recode.stats | cut -f1 --complement >> freebayes.quality.stats 
-```
-
-```{r echo = F}
 fig_quality_dist <- qual_df %>% 
@@
   theme_bw()
-
 ```
 
-\newpage
+### R code to output tables and figures
 
-# Discussion
+```{r supplementary3, eval = F, echo=T, include=T}
+## table 1
+tribble(
+  ~ Caller, ~ `raw SNPs`, ~ `raw ts/tv`, ~ `post-filter SNPs`, ~ `post-filter ts/tv`,
+  "mpileup", 132046, 0.81, 11115, 0.87,
+  "freebayes", 106973, 0.44, 6722, 1.09
+) %>% 
+  pander::pander(caption = "Raw and filtered SNP counts and transition/transversion ratios for calls by mpileup and freebayes.")
 
-<!-- Sites with low quality were discarded -->
+### fig 1
+(fig_allelefreq + fig_sitedepth) &
+  plot_layout(guides = 'keep') &
+  plot_annotation(tag_levels = 'A')
 
-<!-- Kept all allele freq, even rare ones ... rare ones are cool because reference... depends of system of interest though ... -->
+## table 2
+Overlap %>% pander::pander(caption = "Summary of total, shared, and unique SNPs called by mpileup and freebayes (after filtering).")
 
-<!-- Sites with extremely large depth could represent paralogous regions? -->
-
-Comparison of callers: Hwang2015; Cornish2015
-
--   Reflect on results
-
--   Concurrence with other findings
-
-    -   If not, why not (data...)
-
--   Characteristics of the data affect variant calls?
-
-    -   Reference was highly fragmented, though \~85 % reads mapped to the reference in all individuals (in earlier work)
-
-    -
-
--   Other features of the VCF that are notable ?
-The mean read depth across all sites is larger for all individuals in the `freebayes` SNPs (fig.
-2).
-Perhaps this is because freebayes is less tolerant of low-coverage sites in it's estimation of genotype likelihood/ quality / ....
-
--   What other filters:
-
-    -   Depth.... high coverage can indicate a paralogous region (duplicated)
-
-    -   Might be interested in triallelic snps or indels, depending on the study...
-
-    -   If looking for rare variants we would perhaps want to not use filters and only look at positions that are relevant to the process or disease of interest [@cornish2015].
-
-Both `freebayes` and `mpileup` apply Bayes' theorem in their algorithms for SNP calling.
-
-mpileup considers the bases at a variant position
-
-freebayes uses a window approach that differs from other bayesian callers which treat the alignment similarly...
-
-cornish
-
--   callers have high positive predictive value (TP/TP+FP) but low sensitivity (TP/P).
-
--   Freebayes identified more raw SNPs (far more false positive) in the human exome data; but mpileup and freebayes had similar counts after filtering.
-
--   Freebayes did not perform as well as mpileup.
-
-------------------------------------------------------------------------
-
-# Supplemental Figures
-
-```{r}
-fig_alleledepth_over_maf
-```
-
-```{r fig2, echo=F, fig.align = 'center', fig.width=5, fig.height=2.9, fig.cap="Mean depth per locus for each individual."}
-fig_indvdepth
-```
-
-```{r fig3, echo=F, fig.align = 'center', fig.width=5, fig.height=2.9, fig.cap="Mean depth per locus for each individual."}
-fig_alleledepth_over_maf
-```
-
-```{r}
-
+### fig 2
+(fig_alleledepth_over_maf / fig_indvdepth / fig_quality_dist) & 
+  plot_annotation(tag_levels = 'A')
 ```

### 2021-03-19 manuscript.Rmd
almost finished anyway

@@
   pdf_document:
+    highlight: kate
+  html_notebook:
     highlight: kate
@@
 Jason Moggridge 1159229  
+2021/04/19
 
 ```{r setup, echo=F}
-knitr::opts_chunk$set(warning = F, message = F)
+knitr::opts_chunk$set(warning = F, message = F, cache = T)
 ```
@@
 Both `bcftools` and `freebayes` apply Bayesian models to estimate genotype posterior probabilities, though they differ in how this is implemented.
-`bcftools` uses a two step process where `mpileup` first creates the 'pileup' of all reads at variant sites, and then `call` is used to infer genotypes at these [@li2011, @danecek2021].
+`bcftools` uses a two step process where `mpileup` first creates the 'pileup' of all reads at variant sites, and then `call` is used to infer genotypes at these [@li2011; @danecek2021].
 This implementation has the advantage of being very fast.
 `freebayes` a uses different approach to inferring haplotypes in that it does not use the alignment but the full reads [@garrison2012; @howtof].
-Previous comparisons of variant callers have often found low concordance between tools and issues with sensitivity [@orawe2013, @cornish2015].
-However, others have reported high concordance and accuracy in their evaluation [@hwang2015]; the best method is likely dependent on nature of the sequencing data: (whole-genome or exome, RAD-seq, RNA-seq, *etc*.), the depth of coverage, and the type of variants of interest (*eg.* SNPs or indels).
+Previous comparisons of variant callers have often found low concordance between tools and issues with sensitivity [@orawe2013; @cornish2015].
+However, others have reported high concordance and accuracy in their evaluation [@hwang2015]; the best method is likely dependent on nature of the sequencing data (*eg.* whole-genome or exome, RAD-seq, RNA-seq, *etc*.), the depth of coverage, and the type of variants of interest (*eg.* SNPs or indels).
 
@@
 
-```{r load data, message=FALSE}
+```{r load data, message=FALSE, echo=FALSE}
 #! {R code}
@@
   facet_wrap(caller~., scales = 'free_y') +
-  labs(x = 'minor allele frequency', y='loci') +
+  labs(x = 'Minor Allele Frequency', y='Number of Loci') +
   theme_bw()
@@
   scale_x_log10() +
-  labs(x = 'depth per locus', y = 'density') +
+  labs(x = 'Depth', y = 'Density') +
   guides(colour = guide_legend(override.aes = list(size=1, pch=15))) +
@@
   labs(x = "Minor allele frequency",
-       y = "Allele depth", 
+       y = "Depth", 
        color="Number of\nchromosomes") +
@@
   facet_wrap( ~ caller, scales = 'free_x') +
-  labs(y = 'mean depth per locus', x = '', color = 'Individual') +
+  labs(y = 'Mean Depth', x = '', color = 'Subject') +
   theme_bw() +
@@
   scale_fill_carto_d(palette = 2) +
-  labs(x="quality score", y = "loci", fill = NULL) +
+  labs(x="Quality Score", y = "Number of Loci", fill = NULL) +
   scale_y_log10() +
@@
 
-\newpage
-
 # Results
 
-From the ten burbots in our sample, `mpileup` called a total of 132,046 SNPs, 173 of which were multiallelic, and with a transition:transversion ratio (ts:tv) of 0.81; `freebayes`called a total of 106,973 SNPs, with 1761 being multiallelic, and ts:tv of 0.44 (table 1).
-After applying filters for these were reduced to 11,115 SNPs with a 0.87 transition:transversion ratio for `mpileup` calls and 6,722 SNPs with 1.09 transition:transversion ratio for `freebayes` calls.
-The sets of biallelic SNPs identified by each caller were compared (table 2): `mpileup` called `r round(Overlap[1,2][[1]] * 100 / Overlap[2,2][[1]], 2) - 100` % more biallelic SNPs than `freebayes`.
-Of these, a large proportion (`r scales::comma(Overlap[3,2][[1]])`) were shared between both sets, representing `r round(Overlap[3,2][[1]] * 100 /Overlap[1,2][[1]], 2)` % of `mpileup` calls and `r round(Overlap[3,2][[1]] * 100 /Overlap[2,2][[1]], 2)` % of `freebayes` calls.
-Additionally, `r Overlap[4,2][[1]]` SNPs were unique to `mpileup` and `r Overlap[5,2][[1]]` to `freebayes`.
-
-The distribution minor allele frequencies of SNPs from `mpileup` shows a U-shape with peaks at 0.05 and 0.5, whereas those from `freebayes` peak at 0.05 and are otherwise evenly distributed (fig.1A).
-The range of possible values is constrained due to the small sample size (some values can only be obtained with a specific amount of missing data; fig.
-2A).
-The sum depth per locus across all individuals is generally in the range of 50-200 reads, with both callers showing similar distributions (fig.
-1B).
-The mean depth for `mpileup` was 173.7 reads and for `freebayes` 162.2 reads.
-Both callers have some extreme outliers in terms of site depth: the maximum depth for `freebayes` was 16,162 reads, while for `mpileup` it was only 2522 reads.
-The minimum site depth was 5 and 3 for `freebayes` and `mpileup` respectively and `mpileup` has a greater proportion of low-depth sites in general.
-Freebayes also had a greater mean depth per individual (fig.
-2B).
-Both callers show similar distributions of quality scores, though `mpileup` calls a greater proportion of transversions at low-quality sites (fig.
-2C).
-
-------------------------------------------------------------------------
+From the ten burbots in our sample, mpileup called a total of 132,046 SNPs, 173 of which were multiallelic, and with a transition:transversion ratio (ts:tv) of 0.81; freebayes called a total of 106,973 SNPs, with 1761 being multiallelic, and ts:tv of 0.44 (table 1).
+After applying filters for these were reduced to 11,115 SNPs with a 0.87 transition:transversion ratio for mpileup calls and 6,722 SNPs with 1.09 transition:transversion ratio for freebayes calls.
+The sets of biallelic SNPs identified by each caller were compared (table 2): mpileup called `r round(Overlap[1,2][[1]]* 100 / Overlap[2,2][[1]], 2) - 100` % more biallelic SNPs than freebayes.
+Of these, a large proportion (`r scales::comma(Overlap[3,2][[1]])`) were shared between both sets, representing `r round(Overlap[3,2][[1]]*100 /Overlap[1,2][[1]], 2)` % of mpileup calls and `r round(Overlap[3,2][[1]]*100 /Overlap[2,2][[1]], 2)` % of freebayes calls.
+Additionally, `r Overlap[4,2][[1]]` SNPs were unique to mpileup and `r Overlap[5,2][[1]]` unique to freebayes.
 
@@
   ~ Caller, ~ `raw SNPs`, ~ `raw ts/tv`, ~ `post-filter SNPs`, ~ `post-filter ts/tv`,
-  "mpileup", 132046, 0.81, 11115, 0.87,
-  "freebayes", 106973, 0.44, 6722, 1.09
-) %>% pander::pander(caption = "Raw and filtered SNP counts and transition/transversion ratios for calls by mpileup and freebayes.")
+  "mpileup", '132,046', 0.81, '11,115', 0.87,
+  "freebayes", '106,973', 0.44, '6,722', 1.09
+) %>% 
+  pander::pander(caption = "Raw and filtered SNP counts and transition/transversion ratios for calls by mpileup and freebayes.")
 ```
 
-```{r fig1, echo=F, fig.align = 'center', fig.width=7, fig.height=3, fig.cap="(A) Distribution of minor allele frequency; SNPs with frequency = 0 or 1 were omitted.
-(B) Distribution of depth per locus summed across all individuals (note log-scaled x-axis)."}
+  The distribution minor allele frequencies of SNPs from mpileup shows a U-shape with peaks at 0.05 and 0.5, whereas those from freebayes peak at 0.05 and are otherwise evenly distributed (fig.1A).
+The range of possible values is constrained due to the small sample size (some values can only be obtained with a specific amount of missing data; fig.
+2A).
+The sum depth per locus across all individuals is generally in the range of 75x-200x, with both callers showing similar distributions (fig.
+1B).
+The mean depth for mpileup was 174 reads and for freebayes 162 reads.
+Both callers have some extreme outliers in terms of site depth: the maximum depth for freebayes was 16,162 reads, while for mpileup it was only 2522 reads.
+The minimum site depth was 5 and 3 for freebayes and mpileup respectively and mpileup has a greater proportion of low-depth sites in general.
+Freebayes also had a greater mean depth per individual (fig.
+2B).
+Both callers show similar distributions of quality scores, though mpileup calls a greater proportion of transversions at low-quality loci than freebayes (fig.
+2C).
+
+-----
+
+```{r table2_overlap, echo=F}
+Overlap %>% 
+  mutate(Count = scales::comma(Count)) %>% 
+  pander::pander(caption = "Summary of total, shared, and unique SNPs called by mpileup and freebayes (after filtering).")
+```
+
+```{r fig1, echo=F, fig.align = 'center', fig.width=8, fig.height=3, fig.cap="(A) Distribution of minor allele frequency; SNPs with frequency = 0 or 1 were omitted.
+(B) Distribution of depth per locus summed across all individuals (nb: the x-axis is log-scaled)."}
 (fig_allelefreq + fig_sitedepth) &
@@
 
-```{r table2_overlap, echo=F}
-Overlap %>% 
-  pander::pander(caption = "Summary of total, shared, and unique SNPs called by mpileup and freebayes (after filtering).")
-```
-
-```{r fig2, echo=F, fig.align = 'center', fig.width=6, fig.height=7, fig.cap="(A) Scatterplot of loci (points) by allele depth and minor allele frequency, colored by the number of chromosomes counted.
-Due to the sample size (n=10), minor allele frequencies are constrained by the number of chromosomes at the SNP site; there appears to be no relationship between allele depth (ie.
+```{r fig2, echo=F, fig.align = 'center', fig.width=6, fig.height=8, fig.cap="(A) Scatterplot of loci (points) by allele depth and minor allele frequency, colored by the number of chromosomes counted.
+Due to the sample size (n=10), minor allele frequencies are constrained by the number of chromosomes at the SNP site; there appears to be no relationship between depth (ie.
 read coverage) and minor allele frequencies.
@@
 ```
+
+----
 
@@
 The highly-fragmented assembly means we are likely to miss many variants since they are not represented in the alignment.
-The relatively low coverage of the data means that a number of true variants are likely discarded because of high uncertainty in their classification.
-Further, burbot is not a model organism and does not have the same resources to develop useful prior probabilities for variant calling (eg.
-SNP databases for human variant calling).
+The relatively low coverage of the data means that a number of true variants are likely discarded because of large uncertainty in their classification.
+Further, burbot is not a model organism and does not have the same resources to develop useful prior probabilities for variant calling, *eg*.
+SNP databases for human variant calling [@nielsen2011].
+
 In the present work, the small number of subjects means that the identification of uncommon and rare variants is highly unlikely.
@@
 
-I also examined the quality score distribution over sites and the mean per site depth by individual. These were computed with `bcftools stats` and `vcftools` respectively, as follows:
+I also examined the quality score distribution over sites and the mean per site depth by individual. These were computed with `bcftools stats` and `vcftools` respectively, as follows (bash code):
 
-```{r engine = 'bash', eval=FALSE}
+```{r stats_quality, engine = 'bash', eval=FALSE}
 ## quality scores from bf
@@
 ```{r supplementary2, eval = F, echo=T, include=T}
+## Figure 1L
 # Allele frequency histogram (alleles stacked)
@@
 
-## SUPPLEMENTAL
+## Figure 2:
 
@@
   )
+# Quality score distribution; fill bars with type: ts or tv 
 fig_quality_dist <- qual_df %>% 
@@
   ) %>% 
-  pander::pander(caption = "Raw and filtered SNP counts and transition/transversion ratios for calls by mpileup and freebayes.")
-
+  pander::pander(caption = "Raw and filtered SNP counts and
+                 transition/transversion ratios for calls by mpileup 
+                 and freebayes.")
 ### fig 1
@@
   plot_annotation(tag_levels = 'A')
-
 ## table 2
-Overlap %>% pander::pander(caption = "Summary of total, shared, and unique SNPs called by mpileup and freebayes (after filtering).")
+Overlap %>% 
+  pander::pander(caption = "Summary of total, shared, and unique SNPs
+                 called by mpileup and freebayes (after filtering).")
 

### 2021-03-19 manuscript.Rmd
done for the night; figures look fine; need to finish last paragraph about filter choice

@@
 ---
-title: ''
-author: ""
+title: "Project 3: Comparison of variant calling by `freebayes` and `mpileup` from high-throughput sequencing of Burbot (*Lota lota*)"
+author: "Jason Moggridge 1159229"
+date: "2021/04/19"
 output: 
@@
 
-# Project 3: Comparison of variant calling by `freebayes` and `mpileup` from high-throughput sequencing of Burbot (*Lota lota*)
-
-Jason Moggridge 1159229  
-2021/04/19
-
 ```{r setup, echo=F}
@@
 
-The analysis of genetic variants (single-nucleotide polymorphisms -SNPs and insertions/deletions - indels) by high-throughput sequencing has applications in studies of phylogeny and population structure, quantitative-trait loci, genome-wide associations, among others.
-Such polymorphic sites are also of practical interest in clinical genomics, *eg*.
+The analysis of genetic variants such as single-nucleotide polymorphisms (SNPs) and insertions/deletions (indels) by high-throughput sequencing is a central objective in studies of quantitative-trait loci, genome-wide associations, population structure, among others.
+These polymorphic sites are also of practical interest in clinical genomics, *eg*.
 to discover disease-causing mutations [@cornish2015], and in agriculture, where SNPs can be used as genetic markers to guide breeding of crops [@yao2020] and livestock [@ni2015].
@@
 The identification of polymorphic sites from sequencing data, known as variant calling, is performed after alignment of reads to a reference genome.
-Because high-throughput sequencing and read-mapping are error-prone, it is challenging but essential to differentiate true variants from artefacts, particularly in low-coverage data [@nielsen2011].
-Many tools for variant calling have been created for this classification task (`bcftools`, `GATK`, `freebayes`, *etc.*), though these differ in their statistical models for inferring genotypes.
-Further filtering of variants is generally applied after calling, and how this is performed will vary depending on the nature of the study.
+Because high-throughput sequencing and read-mapping are noisy, it is challenging but essential to differentiate true variants from artefacts, particularly in low-coverage data [@nielsen2011].
+Many variant calling tools (`bcftools`, `GATK`, `freebayes`, *etc.*) have been created for this classification task, though these differ in their error correction processes and statistical models for inferring genotypes.
+Further filtering of variants is generally applied after calling, and how this is performed depends on the nature of the study.
 
@@
 IndvDepth <- bind_rows(
+    read_tsv('./data/freebayes.idepth', col_names = cnames, skip = 1L) %>%
+      mutate(caller = 'freebayes'),
     read_tsv('./data/mpileup.idepth', col_names = cnames, skip = 1L) %>%
-      mutate(caller = 'mpileup'),
-    read_tsv('./data/freebayes.idepth', col_names = cnames, skip = 1L) %>%
-      mutate(caller = 'freebayes')
+      mutate(caller = 'mpileup')
   ) %>% 
@@
 SiteDepth <- bind_rows(
+  read_tsv('./data/freebayes.ldepth', col_names = cnames, skip = 1L) %>%
+    mutate(caller = 'freebayes'),
   read_tsv('./data/mpileup.ldepth', col_names = cnames, skip = 1L) %>%
-    mutate(caller = 'mpileup'),
-  read_tsv('./data/freebayes.ldepth', col_names = cnames, skip = 1L) %>%
-    mutate(caller = 'freebayes')
+    mutate(caller = 'mpileup')
 ) %>% mutate(caller = as_factor(caller))
@@
   geom_histogram(aes(x = maf), bins = 20, alpha = 0.8) +
-  facet_wrap(caller~., scales = 'free_y') +
+  facet_wrap(~caller, scales = 'free_y', nrow = 2) +
   labs(x = 'Minor Allele Frequency', y='Number of Loci') +
+  theme_bw()
+
+# Allele depth over maf, coloured by n chromosomes
+fig_alleledepth_over_maf <- 
+  ggplot(vcf_df, aes(x=maf, y=allele_depth, color = n_chr)) +
+  geom_point(size = 0.5, alpha = 0.5) +
+  scale_y_log10() +
+  facet_wrap(caller~., scales = 'free_y', nrow = 2) +
+  scale_color_viridis_c() +
+  labs(x = "Minor allele frequency",
+       y = "Depth", 
+       color="Number of\nchromosomes") +
   theme_bw()
@@
   guides(colour = guide_legend(override.aes = list(size=1, pch=15))) +
-  theme_bw() +
-  theme(legend.position = c(.8,.7))
-
-## SUPPLEMENTAL
-
-# Allele depth over maf, coloured by n chromosomes
-fig_alleledepth_over_maf <- 
-  ggplot(vcf_df, aes(x=maf, y=allele_depth, color = n_chr)) +
-  geom_point(size = 0.5, alpha = 0.7) +
-  scale_y_log10() +
-  facet_wrap(~caller) +
-  scale_color_viridis_c() +
-  labs(x = "Minor allele frequency",
-       y = "Depth", 
-       color="Number of\nchromosomes") +
   theme_bw()
+  # theme(legend.position = c(.8,.7))
 
@@
 The range of possible values is constrained due to the small sample size (some values can only be obtained with a specific amount of missing data; fig.
+1B).
+The sum depth per locus across all individuals is generally in the range of 75x-200x, with both callers showing similar distributions (fig.
 2A).
-The sum depth per locus across all individuals is generally in the range of 75x-200x, with both callers showing similar distributions (fig.
-1B).
 The mean depth for mpileup was 174 reads and for freebayes 162 reads.
@@
 
------
-
 ```{r table2_overlap, echo=F}
@@
 
-```{r fig1, echo=F, fig.align = 'center', fig.width=8, fig.height=3, fig.cap="(A) Distribution of minor allele frequency; SNPs with frequency = 0 or 1 were omitted.
+```{r fig1, echo=F, fig.align = 'center', fig.width=8, fig.height=4, fig.cap="(A) Distribution of minor allele frequency; SNPs with frequency = 0 or 1 were omitted.
 (B) Distribution of depth per locus summed across all individuals (nb: the x-axis is log-scaled)."}
-(fig_allelefreq + fig_sitedepth) &
+(fig_allelefreq + fig_alleledepth_over_maf) &
   plot_layout(guides = 'keep') &
@@
 \n(C) The quality score distributions for each set of SNP calls, separated into transitons (ts) and transversions (tv)"}
-fig_alleledepth_over_maf / fig_indvdepth / fig_quality_dist & 
+ fig_sitedepth / fig_indvdepth / fig_quality_dist & 
   plot_annotation(tag_levels = 'A')
@@
 
-----
+------------------------------------------------------------------------
 
@@
 Both callers have a peak at low MAF (0.05) but otherwise mpileup has an even distribution from 0.10 to 0.5, when accounting for the constraint on these values by the small sample size (fig.
-2A).
+1B).
 The callers show differences in the read depth across loci (fig.
-1B) and individuals (fig.
-2B), as high-depth sites (\>250/ind.) are hard-filtered by `mpileup`, and mpileup identifies many more low-depth sites (\<70x) that are ignored by freebayes.
+2A) and individuals (fig.
+2B), as high-depth sites (\>250x/ind.) are hard-filtered by `mpileup`, and mpileup identifies many more low-depth sites (\<70x) that are ignored by freebayes.
 As such, the concordance between sets would likely increase if stricter filters were used to mitigate these biases.
@@
 The relatively low coverage of the data means that a number of true variants are likely discarded because of large uncertainty in their classification.
-Further, burbot is not a model organism and does not have the same resources to develop useful prior probabilities for variant calling, *eg*.
-SNP databases for human variant calling [@nielsen2011].
+Further, burbot is not a model organism and does not have the same resources to develop useful prior probabilities for variant calling, (*eg*.
+SNP databases for human variant calling; [@nielsen2011]), or benchmark data to validate pipelines (such as GiaB; [@zook2019a]).
 
@@
 If we were more interested in finding such loci, we would want to apply less stringent filtering and accept the cost of increased false discoveries.
+However,
 
@@
   labs(x = 'depth per locus', y = 'density') +
-  guides(colour = guide_legend(override.aes = list(size=1, pch=15))) +
-  theme_bw() +
-  theme(legend.position = c(.8,.7))
+  guides(colour = guide_legend(override.aes = list(size=0.5, pch=15))) +
+  theme_bw()
 
@@
   scale_color_carto_d(palette = 2) +
-  guides(colour = guide_legend(override.aes = list(size = 4, pch = 16))) +
+  guides(colour = guide_legend(override.aes = list(size = 3, pch = 16))) +
   facet_wrap( ~ caller, scales = 'free_x') +
@@
   theme_bw() +
-  theme(
-    axis.text.x = element_blank(),
+  theme(axis.text.x = element_blank(),
         axis.ticks.x = element_blank(),
-    legend.position = 'right'
-  )
+        legend.position = 'right')
+
 # Quality score distribution; fill bars with type: ts or tv 
@@
   facet_wrap(~caller) +
-  scale_fill_carto_d(palette = 2) +
+  scale_fill_carto_d(palette = 1, direction = -1) +
   labs(x="quality score", y = "loci", fill = NULL) +

### 2021-03-19 manuscript.Rmd
proofreading time

@@
 fig_sitedepth <- 
-  ggplot(SiteDepth, aes(x = allele_depth, fill = caller, color = caller)) +
+  ggplot(SiteDepth, aes(x = allele_depth/10,
+                        fill = caller, color = caller)) +
   geom_density(alpha = 0.12) +
-  geom_vline(aes(xintercept = mean(allele_depth), color = caller),
-             lty = 2) +
   scale_x_log10() +
-  labs(x = 'Depth', y = 'Density') +
+  labs(x = 'Mean Site Depth per Individual', y = 'Density') +
   guides(colour = guide_legend(override.aes = list(size=1, pch=15))) +
@@
   scale_color_carto_d(palette = 2) +
-  guides(colour = guide_legend(override.aes = list(size = 4, pch = 16))) +
+  guides(colour = guide_legend(override.aes = list(size = 2, pch = 16, keyheight=0.01, default.unit="cm"))) +
   facet_wrap( ~ caller, scales = 'free_x') +
@@
     axis.ticks.x = element_blank(),
-    legend.position = 'right'
+    legend.position = 'right', legend.text = element_text(size=8),
+    legend.spacing.y = unit(0.01, 'cm')
   )
+
 fig_quality_dist <- qual_df %>% 
@@
 1B).
-The sum depth per locus across all individuals is generally in the range of 75x-200x, with both callers showing similar distributions (fig.
+The mean depth per locus across individuals is mostly in the range of 8x-30x, with both callers showing peaks at 25x and similar distributions (fig.
 2A).
-The mean depth for mpileup was 174 reads and for freebayes 162 reads.
+The mean depth (per site per subject) for mpileup was 17x and for freebayes 16x reads.
 Both callers have some extreme outliers in terms of site depth: the maximum depth for freebayes was 16,162 reads, while for mpileup it was only 2522 reads.
@@
 
-```{r fig1, echo=F, fig.align = 'center', fig.width=8, fig.height=4, fig.cap="(A) Distribution of minor allele frequency; SNPs with frequency = 0 or 1 were omitted.
-(B) Distribution of depth per locus summed across all individuals (nb: the x-axis is log-scaled)."}
+```{r fig1, echo=F, fig.align = 'center', fig.width=8, fig.height=4, fig.cap="(A) Distribution of minor allele frequency of freebayes and mpileup SNP calls; SNPs with frequency = 0 or 1 were discarded in filtering (B) Scatterplot of loci (points) by allele depth and minor allele frequency, colored by the number of chromosomes counted.
+Due to the sample size (n=10), minor allele frequencies are constrained by the number of chromosomes at the SNP site; there appears to be no relationship between depth (ie.
+read coverage) and minor allele frequencies.
+"}
 (fig_allelefreq + fig_alleledepth_over_maf) &
@@
 
-```{r fig2, echo=F, fig.align = 'center', fig.width=6, fig.height=8, fig.cap="(A) Scatterplot of loci (points) by allele depth and minor allele frequency, colored by the number of chromosomes counted.
-Due to the sample size (n=10), minor allele frequencies are constrained by the number of chromosomes at the SNP site; there appears to be no relationship between depth (ie.
-read coverage) and minor allele frequencies.
-\n(B) Mean depth per locus for each individual.
+```{r fig2, echo=F, fig.align = 'center', fig.width=6, fig.height=8, fig.cap="(A) Distribution of mean depth at SNP loci identified by freebayes and mpileup from 10 burbot subjects (nb: the x-axis is log-scaled).(B) Mean depth per locus for each individual.
 Violin plot lines indicate quartiles.
@@
 ```
-
-------------------------------------------------------------------------
 
@@
 
-In the present work, the small number of subjects means that the identification of uncommon and rare variants is highly unlikely.
-If we were more interested in finding such loci, we would want to apply less stringent filtering and accept the cost of increased false discoveries.
-However,
+In the present work, the small number of subjects means that the identification of uncommon variants is highly unlikely.
+If we interested in performing a population structure analysis of burbot, we should sequence more subjects and instead apply a filter for minor allele count instead of frequency to incorporate information from uncommon and rare variants [@linck2019].
+Application of additional filters, *eg*.
+for strand-bias, site depth, and depth variation, would likely improve the accuracy of variant calling [@oleary2018].
 

### 2021-03-19 Project3_JMoggridge.Rmd
submitted

--- 2021-03-19 manuscript.Rmd
@@
 ```{r setup, echo=F}
-knitr::opts_chunk$set(warning = F, message = F, cache = T)
+knitr::opts_chunk$set(warning = F, message = F, cache = F)
 ```
@@
 
-The analysis of genetic variants such as single-nucleotide polymorphisms (SNPs) and insertions/deletions (indels) by high-throughput sequencing is a central objective in studies of quantitative-trait loci, genome-wide associations, population structure, among others.
+The analysis of genetic variants such as single-nucleotide polymorphisms (SNPs) and insertions/deletions (indels) by high-throughput sequencing is a central objective in studies of quantitative-trait loci, genome-wide associations, and population structure, among others.
 These polymorphic sites are also of practical interest in clinical genomics, *eg*.
 to discover disease-causing mutations [@cornish2015], and in agriculture, where SNPs can be used as genetic markers to guide breeding of crops [@yao2020] and livestock [@ni2015].
-As the cost of sequencing has decreased greatly and computational resources have improved, these techniques have become increasingly popular.
+As the cost of sequencing has decreased and computational resources have improved, these techniques have become increasingly popular.
 
@@
 
-In the current work I compared variant calling by two tools: `bcftools mpileup/call` (previously `samtools`) and `freebayes`.
-The results were compared in terms of the number of SNPs, the overlap between sets, the depth at each position, and their minor allele frequencies.
-Both `bcftools` and `freebayes` apply Bayesian models to estimate genotype posterior probabilities, though they differ in how this is implemented.
+Both `bcftools mpileup/call` and `freebayes` apply Bayesian models to estimate genotype posterior probabilities, though they differ in how this is implemented.
 `bcftools` uses a two step process where `mpileup` first creates the 'pileup' of all reads at variant sites, and then `call` is used to infer genotypes at these [@li2011; @danecek2021].
@@
 `freebayes` a uses different approach to inferring haplotypes in that it does not use the alignment but the full reads [@garrison2012; @howtof].
+
+I compared variant calling by two tools: `bcftools mpileup/call` and `freebayes`.
+The results were compared in terms of the number of SNPs, the overlap between sets, the depth at each position, and their minor allele frequencies.
 Previous comparisons of variant callers have often found low concordance between tools and issues with sensitivity [@orawe2013; @cornish2015].
-However, others have reported high concordance and accuracy in their evaluation [@hwang2015]; the best method is likely dependent on nature of the sequencing data (*eg.* whole-genome or exome, RAD-seq, RNA-seq, *etc*.), the depth of coverage, and the type of variants of interest (*eg.* SNPs or indels).
+However, others have reported high concordance and accuracy in their evaluation [@hwang2015]; the best method is likely dependent on nature of the sequencing data (*eg.* whole-genome or exome, RAD-seq, RNA-seq, *etc*.), the sequencing depth, and the type of variants of interest (*eg.* SNPs or indels).
 
@@
   bind_rows(
-    read_tsv("./data/mpileup.quality.stats", skip = 1L, col_names = cnames) %>%  mutate(caller = "mpileup"),
-    read_tsv("./data/freebayes.quality.stats", skip = 1L, col_names = cnames) %>%  mutate(caller = "freebayes")
+    read_tsv("./data/freebayes.quality.stats", skip = 1L, col_names = cnames) %>%  mutate(caller = "freebayes"),
+    read_tsv("./data/mpileup.quality.stats", skip = 1L, col_names = cnames) %>%  mutate(caller = "mpileup")
   ) %>%  mutate(caller = as_factor(caller))
@@
 2A).
-The mean depth (per site per subject) for mpileup was 17x and for freebayes 16x reads.
-Both callers have some extreme outliers in terms of site depth: the maximum depth for freebayes was 16,162 reads, while for mpileup it was only 2522 reads.
-The minimum site depth was 5 and 3 for freebayes and mpileup respectively and mpileup has a greater proportion of low-depth sites in general.
-Freebayes also had a greater mean depth per individual (fig.
+The mean depth (per site, per subject) for mpileup was 17x and for freebayes 16x reads.
+Both callers have some extreme outliers in terms of site depth: the maximum depth (summed across individuals) for freebayes was 16,162 reads, while for mpileup it was only 2522 reads.
+The minimum site depth (summed across individuals) was 5 and 3 for freebayes and mpileup respectively and mpileup calls a greater proportion of low-depth sites in general (fig.
+2A).
+Freebayes also had a greater mean depth for each individual (fig.
 2B).
-Both callers show similar distributions of quality scores, though mpileup calls a greater proportion of transversions at low-quality loci than freebayes (fig.
+Both tools show similar distributions of quality scores, though mpileup calls a greater proportion of transversions at low-quality loci than freebayes (fig.
 2C).
@@
 2A) and individuals (fig.
-2B), as high-depth sites (\>250x/ind.) are hard-filtered by `mpileup`, and mpileup identifies many more low-depth sites (\<70x) that are ignored by freebayes.
+2B), as high-depth sites (\>250x/subject) are hard-filtered by `mpileup`, and mpileup identifies many more low-depth sites (mean \< 7x/subject) that are ignored by freebayes.
 As such, the concordance between sets would likely increase if stricter filters were used to mitigate these biases.
@@
 The relatively low coverage of the data means that a number of true variants are likely discarded because of large uncertainty in their classification.
-Further, burbot is not a model organism and does not have the same resources to develop useful prior probabilities for variant calling, (*eg*.
-SNP databases for human variant calling; [@nielsen2011]), or benchmark data to validate pipelines (such as GiaB; [@zook2019a]).
+Further, burbot is not a model organism and does not have the same resources to develop useful prior probabilities for variant calling, *eg*.
+SNP databases for human variant calling [@nielsen2011], or benchmark datasets (such as GiaB) to validate pipelines [@zook2019a].
 
@@
 
-### R code for plotting results
+### R code for creating figures
 
